---
title: "Comparing integration method outputs"
author: "Joe Oddy"
lightbox: auto
execute:
  echo: false
  warning: false
  dev: "ragg_png"
  cache: false
format:
  html:
    theme: flatly
    mermaid:
      theme: neutral
    page-layout: article
    self-contained: true
    code-line-numbers: true
    code-block-bg: true
    code-block-border-left: "#30BBF0"
    code-copy: true
    code-link: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    fig-align: center
    fig-height: 5
    fig-width: 8
    fig-responsive: true
    fig-dpi: 600
    fig-cap-location: top
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    dev: "ragg_png"
    smooth-scroll: true
    embed-resources: true
---

# Loading

```{r}
#| label: library-loading

library(here)
library(aqua)
library(umap)
library(mixOmics)
# mixKernel has quite a few dependencies, see mixKernel installation guide
library(mixKernel)
library(conflicted)
library(janitor)
library(gt)
library(ggrepel)
library(kernlab)
library(feather)
library(UpSetR)
library(ComplexUpset)
library(arrow)
library(patchwork)
library(tidymodels)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::desc)
conflicts_prefer(aqua::rescale)
conflicts_prefer(ComplexUpset::upset)
conflicts_prefer(arrow::read_feather)
conflicts_prefer(base::as.factor)
conflicted::conflicts_prefer(dplyr::first)
```

```{r}
#| label: sourcing-data-acquisition-script
#| output: FALSE
#| cache-lazy: FALSE

source(here("01_scripts/data_acquisition.R"))

BPPARAM <- BiocParallel::SnowParam(workers = parallel::detectCores() - 1)
```

```{r}
#| label: scRNA-data-loading

var_features <- setdiff(names(combined_run3_run4_variable_genes_union_df)[-c(1:47)],
                              c("plasticell_sample_id", "label", "omics", "site"))

scRNA_data_long <- combined_pseudobulk_df |>
  select(label, omics, all_of(var_features)) |>
  pivot_longer(
    cols = all_of(var_features),
    names_to = "gene",
    values_to = "normalised_expression"
  ) |>
  group_by(gene) |>
  mutate(scaled_expression = scale(normalised_expression)) |>
  mutate(
    vessel = as.numeric(str_extract(label, "(?<=v)\\d+")),
    day = as.numeric(str_extract(label, "(?<=d)\\d+")),
    run = as.numeric(str_extract(label, "(?<=r)\\d+"))
  ) |>
  relocate(run, vessel, day)

scRNA_data_wide <- scRNA_data_long |>
  select(-normalised_expression) |>
  pivot_wider(
    names_from = gene,
    values_from = scaled_expression
  )
```

```{r}
#| label: lcms-data-loading

lcms_data <- processed_lcms$both_runs_data_wide |>
  mutate(omics = "lcms") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )
```

```{r}
#| label: olink-data-loading

long_olink <- processed_olink$data |>
  filter(!Timepoint == "Fresh Media") |>
  filter(DF == "Neat") |>
  select(label, omics, NPX, Assay, vessel, stream, Timepoint, DF) |>
  filter(Assay %in% processed_olink$m_assays) |>
  group_by(Assay) |>
  mutate(scaled_NPX = scale(NPX)) |>
  ungroup()

wide_olink <- long_olink |>
  select(-NPX) |>
  pivot_wider(
    names_from = Assay,
    values_from = scaled_NPX,
    values_fn = mean
  ) |>
  mutate(
    day = as.numeric(str_remove(Timepoint, "Day ")),
    vessel = as.numeric(str_remove(vessel, "V")),
    run = 4
  ) |>
  relocate(run, day)
```


```{r}
#| label: miRNA-data-loading

miRNA_data <- processed_miRNA$processed_miRNA |>
  mutate(omics = "miRNA") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )

processed_miRNA_long <- miRNA_data |>
  pivot_longer(
    cols = starts_with("ENS"),
    names_to = "miRNA_name",
    values_to = "deseq_normalised_and_vst_expression"
  ) |>
  group_by(miRNA_name) |>
  mutate(scaled_expression = scale(deseq_normalised_and_vst_expression)) |>
  ungroup()

miRNA_data_scaled <- processed_miRNA_long |>
  select(-deseq_normalised_and_vst_expression) |>
  pivot_wider(
    names_from = miRNA_name,
    values_from = scaled_expression
  )
```

```{r}
#| label: process-data-loading

process_data <- processed_process_data$all_process_data_unscaled_added_metrics
```

```{r}
#| label: combining-data

combined_data <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-no-missing-data

common_samples <- combined_data |>
  drop_na() |>
  select(label)

lcms_matrix <- lcms_data |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix <- miRNA_data_scaled |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix <- wide_olink |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix <- scRNA_data_wide |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list <- list(
  lcms = lcms_matrix,
  miRNA = miRNA_matrix,
  olink = olink_matrix,
  scRNA = scRNA_matrix
)

stage <- combined_metadata |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  pull(stage)
```

```{r}
#| label: combining-data-and-process-data

combined_data_and_process <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  ) |>
  full_join(
    process_data |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata_and_process <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    process_data |>
      rename(process = "omics") |>
      select(process, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    ifelse(!is.na(process), process, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-with-missing-data
#| cache-lazy: false

summary_metric_samples <- combined_data_and_process |>
  filter(!is.na(cancer_cells_killed_per_PSC)) |>
  select(label)

lcms_matrix_process <- lcms_data |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix_process <- miRNA_data_scaled |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix_process <- wide_olink |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix_process <- scRNA_data_wide |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list_process <- list(
  lcms = lcms_matrix_process,
  miRNA = miRNA_matrix_process,
  olink = olink_matrix_process,
  scRNA = scRNA_matrix_process
)

summary_metric <- process_data |>
  filter(label %in% summary_metric_samples$label) |>
  arrange(label) |>
  select(cancer_cells_killed_per_PSC, label) |>
  column_to_rownames("label") |>
  as.matrix()
```

```{r}
#| label: loading-nk-cell-proportion-data

nk_cells_proportions <- read.csv(here("00_data/nk_cell_proportions_run_4.csv")) |>
  mutate(
    vessel_day = as.character(vessel_day),
    vessel = as.numeric(sub("V(\\d+)_D\\d+", "\\1", vessel_day)),
    day = as.numeric(sub("V\\d+_D(\\d+)", "\\1", vessel_day)),
    label = paste0("r4_v", vessel, "_d", day)
  )

nk_cells_proportions_subset <- nk_cells_proportions |>
  filter(label %in% common_samples$label) |>
  arrange(label)

nk_cells_target <- nk_cells_proportions_subset |>
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()
```

```{r}
#| label: nk-proportions-conversion-to-categorical

nk_cells_proportions_category_data <- nk_cells_proportions |>
  mutate(proportion_NK_category = case_when(
    proportion_NK <= 0.25 ~ "< 25% NKs",
    proportion_NK > 0.25 & proportion_NK <= 0.5 ~ "25 - 50% NKs",
    proportion_NK > 0.5 & proportion_NK <= 0.75 ~ "50 - 75% NKs",
    proportion_NK > 0.75 & proportion_NK <= 1 ~ "> 75% NKs",
  )) |>
  mutate(proportion_NK_category = fct_relevel(
    proportion_NK_category,
    "< 25% NKs",
    "25 - 50% NKs",
    "50 - 75% NKs",
    "> 75% NKs"
  )) |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  select(-X)

nk_cells_proportions_category <- nk_cells_proportions_category_data |>
  column_to_rownames("label") |>
  pull(proportion_NK_category)
```

```{r}
list2env(processed_process_data, envir = .GlobalEnv)
```
# Supplementary figures

## Supplementary figure 1

```{r}
#| label: fig-omics-features

lcms_features <- lcms_data |>
  select(!omics:vessel) |>
  ncol() |>
  as_tibble() |>
  rename(lcms = value)

miRNA_features <- miRNA_data_scaled |>
  select(!omics:run) |>
  ncol() |>
  as_tibble() |>
  rename(miRNA = value)

olink_features <- wide_olink |>
  select(!run:DF) |>
  ncol() |>
  as_tibble() |>
  rename(olink = value)

scRNA_features <- scRNA_data_wide |>
  select(!run:omics) |>
  ncol() |>
  as_tibble() |>
  rename(scRNA = value)

process_features <- all_process_data_unscaled_added_metrics |>
  select(!omics:label) |>
  ncol() |>
  as_tibble() |>
  rename(process = value)

features_barplot <- bind_cols(lcms_features, miRNA_features, olink_features, scRNA_features) |>
  pivot_longer(
    cols = everything(),
    names_to = "omics",
    values_to = "features"
  ) |>
  ggplot(
    aes(
      x = reorder(omics, features), ,
      y = features
    )
  ) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    x = "Data type",
    y = "Features",
  ) +
  # add numbers in bars
  geom_text(
    aes(label = features),
    hjust = -0.1,
    size = 4
  ) +
  ylim(0, 4200)

features_barplot
```

```{r}
#| label: fig-upset-plot
#| fig-width: 8
#| fig-height: 6

upset_sample_data <- combined_metadata_and_process |> 
  select(miRNA, lcms, olink, scRNA) |> 
  filter(!(is.na(miRNA) & is.na(lcms) & is.na(olink) & is.na(scRNA))) |>
  mutate(across(everything(), ~ ifelse(is.na(.), 0, 1)))

upset_plot <- upset(
  upset_sample_data,
  intersect = c("miRNA", "lcms", "olink", "scRNA"),
  name = "Intersections",
  sort_sets = "descending",
  stripes = "white"
)
```


```{r}
#| label: fig-combined-sample-plots
#| fig-width: 6
#| fig-height: 6

patchwork::wrap_elements(upset_plot) / patchwork::wrap_elements(features_barplot) +
  patchwork::plot_layout(heights = c(2,1)) +
  patchwork::plot_annotation(
    tag_levels = "A"
  )

```

# Clustering figure

## mixKernel

```{r}
scRNA_kernel <- compute.kernel(multiomics_list$scRNA,
  kernel.func = "linear", scale = FALSE
)

miRNA_kernel <- compute.kernel(multiomics_list$miRNA,
  kernel.func = "linear", scale = FALSE
)

lcms_kernel <- compute.kernel(multiomics_list$lcms,
  kernel.func = "linear", scale = FALSE
)

olink_kernel <- compute.kernel(multiomics_list$olink,
  kernel.func = "linear", scale = FALSE
)
```

```{r}
meta_kernel <- combine.kernels(
  scRNA_kernel = scRNA_kernel,
  miRNA_kernel = miRNA_kernel,
  lcms_kernel = lcms_kernel,
  olink_kernel = olink_kernel,
  method = "STATIS-UMKL"
)
```

```{r}
kernel_pca_result <- kernel.pca(meta_kernel, ncomp = 10)
```


```{r}
kernel_pca_data <- kernel_pca_result[["x"]] |> 
  as.data.frame() |>
  rownames_to_column("label") |>
  mutate(stage = stage) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = "multiomics") |> 
  mutate(day= as.factor(str_extract(label, "(?<=d)\\d+"))) |> 
  mutate(day = fct_relevel(
    day,
    "3", "6", "8", "12", "15", "26", "35"
  ))

kernel_pca_plot <- ggplot(kernel_pca_data, aes(x = `1`, y = `2`, color = day)) +
  geom_point(
    size = 2.5,
    alpha = 0.8
  ) +
  theme_catapult() +
  labs(
    x = "PC1",
    y = "PC2",
    color = "Day"
  ) +
  theme_catapult() +
  scale_colour_brewer(palette = "Paired") +
  labs(
    title = "Mixed integration",
    subtitle = "Kernel PCA"
  )

kernel_pca_plot
```

## Early integration clustering

```{r}
#| label: fig-umap-calculation-and-plotting
#| fig-cap: UMAP

set.seed(89)

combined_data_umap <- combined_data |>
  column_to_rownames("label") |>
  drop_na() |>
  umap()

combined_data_umap_values <- combined_data_umap$layout |>
  as.data.frame() |>
  rownames_to_column("label") |>
  full_join(combined_metadata |>
    drop_na())

umap_plot <- ggplot(
  combined_data_umap_values,
  aes(
    x = V1,
    y = V2,
    colour = as.factor(day)
  )
) +
  geom_point(
    size = 2.5,
    alpha = 0.8
  ) +
  scale_color_brewer(palette = "Paired") +
  theme_catapult() +
  labs(
    title = "Early integration",
    subtitle = "UMAP",
    color = "Day"
  )

umap_plot
```

## mixOmics sparse rCCA

Should implement this model here.

```{r}
sgcca_object <- wrapper.sgcca(multiomics_list,
  ncomp = 10,
  scale = FALSE,
  mode = "canonical"
)

sgcca_variates <- sgcca_object$variates

long_variates <- imap_dfr(
  sgcca_variates,
  ~ as.data.frame(.x) |> 
    rownames_to_column("sample") |> 
    mutate(block = .y)
) |> 
  # add stage info
  left_join(combined_metadata |> select(label, stage), by = c("sample" = "label")) |> 
  mutate(day= as.factor(str_extract(sample, "(?<=d)\\d+"))) |> 
  mutate(day = fct_relevel(
    day,
    "3", "6", "8", "12", "15", "26", "35"
  ))

# calculate centroids of each sample
centroids <- long_variates |>
  group_by(sample) |>
  mutate(
    centroid_comp1 = mean(comp1),
    centroid_comp2 = mean(comp2)
  ) |>
  slice_head(n = 1) |> 
  ungroup() |> 
  select(sample, centroid_comp1, centroid_comp2, day)
```

Might want to specify number of features to keep from each block.

```{r}
intermediate_unsupervised_plot <- ggplot(centroids, aes(x = centroid_comp1, y = centroid_comp2, color = day)) +
  geom_point(
    size = 2.5,
    alpha = 0.8
  ) +
  theme_catapult() +
  labs(
    x = "Component 1",
    y = "Component 2",
    color = "Day"
  ) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Intermediate integration",
    subtitle = "Sparse generalised CCA - centroids only",
    color = "Day"
  )

intermediate_unsupervised_plot
```

## Hierarchical integration

Import data from python.

```{r}
#| label: fig-tsne-clustering-plot
#| fig-height: 6
#| fig-width: 6

# read feather file
tsne_data <- arrow::read_feather(here("00_data/tsne_coordinates.feather")) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |> 
  mutate(day= as.factor(str_extract(label, "(?<=d)\\d+"))) |> 
  mutate(day = fct_relevel(
    day,
    "3", "6", "8", "12", "15", "26", "35"
  ))

hierarchical_tsne <- ggplot(tsne_data, aes(x = tSNE1, y = tSNE2, color = day)) +
  geom_point(
    size = 2.5,
    alpha = 0.8
  ) +
  theme_catapult() +
  labs(
    x = "t-SNE 1",
    y = "t-SNE 2",
    color = "Day"
  ) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Hierarchical integration",
    subtitle = "t-SNE"
  )

hierarchical_tsne
```

## Combined plot

```{r}
#| label: fig-combined-clustering-plot
#| fig-height: 8
#| fig-width: 8

(
  (umap_plot | kernel_pca_plot) + plot_layout(widths = c(1, 1))
) /
(
  (intermediate_unsupervised_plot | hierarchical_tsne) + plot_layout(widths = c(1, 1))
) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

# Late integration UMAPs

Similar to the plot above, but with the four panels being the different data types. Probably to go in supplementary.

# Number of important features

Might be good to have a figure similar to the above comparing the top ranked features identified from the different integration methods.

## Late integration features

```{r}
late_integration_feature_importances <- read_csv(here("00_data/feature_importances/late_integration_feature_importances.csv")) |> 
  select(!`...1`)
```

```{r}
late_integration_features_plot <- late_integration_feature_importances |> 
  group_by(omics) |> 
  summarise(n_features = n()) |> 
  ggplot(aes(x = reorder(omics, n_features), y = n_features)) +
  geom_col() +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    hjust = -0.5,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    title = "Late integration",
    subtitle = "glmnet model",
    x = "Omics",
    y = "Number of features"
  )

late_integration_features_plot
```

## Early integration features

```{r}
early_integration_feature_importances <- read_rds(here("00_data/feature_importances/early_integration_feature_importances.rds"))
```

```{r}
#| fig-width: 8
#| fig-height: 3

early_integration_features_plot <- early_integration_feature_importances |> 
  filter(importance > 0) |> 
  group_by(omics) |> 
  summarise(n_features = n()) |> 
  ggplot(aes(x = reorder(omics, n_features), y = n_features)) +
  geom_col() +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    hjust = -0.2,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    title = "Early integration",
    subtitle = "Random forest regression model",
    x = "Omics",
    y = "Number of features"
  )

early_integration_features_plot
  
```

## Intermediate integration

```{r}
intermediate_integration_features <- 
  read_rds(here("00_data/feature_importances/intermediate_integration_features.rds"))

intermediate_integration_features_plot <- intermediate_integration_features |> 
  group_by(omics) |> 
  summarise(n_features = n()) |> 
  ggplot(aes(x = reorder(omics, n_features), y = n_features)) +
  geom_col() +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    hjust = -0.2,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    title = "Intermediate integration",
    subtitle = "MB PLS regression",
    x = "Omics",
    y = "Number of features"
  )

intermediate_integration_features_plot
```


## Hierarchical integration

```{r}
hierarchical_feature_importance <- read_feather(here("00_data/feature_importances/hierarchical_integration_feature_importances.feather")) |> 
  clean_names()
```

```{r}
ggplot(hierarchical_feature_importance,
       aes(x = importance)) +
  geom_histogram() +
  theme_catapult()
```

```{r}
#| label: fig-hierarchical-feature-importance
#| fig-width: 8
#| fig-height: 6

ggplot(hierarchical_feature_importance |> 
         arrange(desc(importance)) |> 
         slice_head(n = 10),
       aes(x = reorder(name, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Hierarchical integration",
    subtitle = "SV pathway importance",
    x = "Feature",
    y = "Importance"
  )
```

```{r}
hierarchical_integration_features_plot <- hierarchical_feature_importance |> 
  filter(importance != 0) |> 
  summarise(n_features = n()) |> 
  mutate(omics = "All") |> 
  ggplot(aes(x = omics, y = n_features)) +
  geom_col() +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    hjust = -0.2,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    title = "Hierarchical integration",
    subtitle = "Random forest regression",
    x = "Omics",
    y = "Number of features"
  )

hierarchical_integration_features_plot
```
## Neural network importances

```{r}
nn_feature_importance <- read_feather(here("00_data/feature_importances/neural_network_importances.feather")) |> 
  clean_names() |> 
  rename(omics = "branch")
```

```{r}
nn_integration_features_plot <- nn_feature_importance |> 
  arrange(-importance) |> 
  slice_head(prop = 0.01) |> 
  group_by(omics) |> 
  summarise(n_features = n()) |> 
  ggplot(aes(x = omics, y = n_features)) +
  geom_col() +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    hjust = -0.2,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    title = "Intermediate integration",
    subtitle = "Multi-branch neural network",
    x = "Omics",
    y = "Number of features"
  )

nn_integration_features_plot
```


## Combined plot

```{r}
#| label: fig-combined-features-plot
#| fig-height: 8
#| fig-width: 8

(
  (early_integration_features_plot / intermediate_integration_features_plot / hierarchical_integration_features_plot / late_integration_features_plot) + plot_layout(heights = c(1, 1, 0.5, 1))
) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-combined-plot-alternative

all_feature_importances <- bind_rows(
  early_integration_feature_importances |> 
    filter(importance > 0) |> 
    mutate(integration = "Early"),
  intermediate_integration_features |> 
    mutate(integration = "Intermediate (MB-PLS)"),
  hierarchical_feature_importance |> 
    filter(importance > 0) |> 
    mutate(integration = "Hierarchical",
           omics = "All"),
  late_integration_feature_importances |> 
    mutate(integration = "Late"),
  nn_feature_importance |> 
    arrange(-importance) |> 
    slice_head(prop = 0.01) |> 
    mutate(integration = "Intermediate (NN)")
)
```

```{r}
#| label: fig-combined-feature-importance-plot
#| fig-height: 5
#| fig-width: 8

all_feature_importances |> 
  group_by(integration, omics) |> 
  summarise(n_features = n()) |> 
  ggplot(aes(x = reorder(integration, n_features), y = n_features, fill = omics)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    position = position_dodge2(width = 0.9, preserve = "single"),
    hjust = -0.2,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    x = "Integration method",
    y = "# of features exceeding importance threshold"
  ) +
  scale_fill_catapult_d() +
  theme(plot.margin = margin(5, 30, 30, 5))
```




# Important features - details

The above plot shows the number of all "important" features. The next ones should identify the top 10 or so from each integration method.

```{r}
all_feature_importances_cleaned <- bind_rows(
  early_integration_feature_importances |> 
    filter(importance > 0) |> 
    mutate(integration = "Early"),
  intermediate_integration_features |>
    select(!feature_omics) |> 
    select(!comp2:comp3) |> 
    filter(comp1 != 0) |> 
    mutate(integration = "Intermediate (MB-PLS)"),
  hierarchical_feature_importance |> 
    filter(importance > 0) |> 
    select(!index) |> 
    rename(feature = "name") |> 
    mutate(integration = "Hierarchical",
           omics = "All"),
  late_integration_feature_importances |> 
    select(!penalty) |> 
    rename(feature = "term") |> 
    mutate(integration = "Late"),
  nn_feature_importance |>
    arrange(-importance) |> 
    slice_head(prop = 0.01) |> 
    mutate(integration = "Intermediate (NN)")
)
```

## Early integration

```{r}
all_feature_importances_cleaned |> 
  filter(integration == "Early") |> 
  arrange(desc(importance)) |> 
  slice_head(n = 8) |> 
  ggplot(aes(x = reorder(feature, importance), y = importance, fill = omics)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Early integration",
    subtitle = "Random forest regression model",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d() -> early_integration_top_features_plot

early_integration_top_features_plot
```

## Intermediate integration

```{r}
all_feature_importances_cleaned |> 
  filter(integration == "Intermediate") |> 
  arrange(desc(abs(comp1))) |> 
  slice_head(n = 8) |> 
  ggplot(aes(x = reorder(feature, abs(comp1)), y = comp1, fill = omics)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Intermediate integration",
    subtitle = "MB-PLS model",
    x = "Feature",
    y = "Loading on component 1"
  ) +
  scale_fill_catapult_d() -> intermediate_integration_top_features_plot

intermediate_integration_top_features_plot
```

## Hierarchical integration

```{r}
#| label: fig-hierarchical-top-features-plot
#| fig-height: 10
#| fig-width: 6

all_feature_importances_cleaned |> 
  filter(integration == "Hierarchical") |> 
  arrange(desc(importance)) |> 
  slice_head(n = 8) |> 
  ggplot(aes(x = reorder(feature, -importance), y = importance)) +
  geom_col() +
  guides(x =  guide_axis(angle = 90)) +
  theme_catapult() +
  labs(
    title = "Hierarchical integration",
    subtitle = "Random forest regression model",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d() -> hierarchical_integration_top_features_plot

hierarchical_integration_top_features_plot
```

## Late integration

```{r}
all_feature_importances_cleaned |> 
  filter(integration == "Late") |> 
  arrange(desc(abs(estimate))) |> 
  slice_head(n = 8) |> 
  ggplot(aes(x = reorder(feature, abs(estimate)), y = estimate, fill = omics)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Late integration",
    subtitle = "Gaussian multiple linear regression with LASSO regularisation",
    x = "Feature",
    y = "Estimate"
  ) +
  scale_fill_catapult_d() -> late_integration_top_features_plot

late_integration_top_features_plot
```

## Late integration faceted

```{r}
#| fig-height: 10
#| fig-width: 8

all_feature_importances_cleaned |> 
  filter(integration == "Late") |> 
  group_by(omics) |> 
  arrange(desc(abs(estimate))) |> 
  slice_head(n = 8) |> 
  ungroup() |> 
  ggplot(aes(x = reorder(feature, abs(estimate)), y = estimate)) +
  facet_wrap(~ omics, scales = "free", ncol = 1) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Late integration",
    subtitle = "Gaussian multiple linear regression with LASSO regularisation",
    x = "Feature",
    y = "Estimate"
  ) +
  scale_fill_catapult_d() -> late_integration_top_features_plot_faceted

late_integration_top_features_plot_faceted
```

## Neural network feature importances

```{r}
#| fig-height: 10
#| fig-width: 8

# nn_feature_importance |> 
#   group_by(branch) |> 
#   arrange(desc(abs(importance))) |> 
#   slice_head(n = 8) |> 
#   ungroup() |> 
#   ggplot(aes(x = reorder(feature, abs(importance)), y = importance)) +
#   facet_wrap(~ branch, scales = "free", ncol = 1) +
#   geom_col() +
#   coord_flip() +
#   theme_catapult() +
#   labs(
#     title = "Intermediate integration",
#     subtitle = "Multi-branch neural network",
#     x = "Feature",
#     y = "Estimate"
#   ) +
#   scale_fill_catapult_d() -> network_top_features_plot_faceted
# 
# network_top_features_plot_faceted
```



## Combined plots

### Hierarchical alone

```{r}
#| label: fig-hierarchical-top-features-plot-alternate
#| fig-height: 4
#| fig-width: 10

all_feature_importances_cleaned |> 
  filter(integration == "Hierarchical") |> 
  arrange(desc(importance)) |> 
  slice_head(n = 8) |> 
  ggplot(aes(x = reorder(feature, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Hierarchical integration",
    subtitle = "Random forest regression model",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d() -> hierarchical_integration_top_features_horiz

hierarchical_integration_top_features_horiz
```


### Without hierarchical

```{r}
#| label: fig-combined-top-features-plot-without
#| fig-height: 14
#| fig-width: 8

(
  (early_integration_top_features_plot / 
     intermediate_integration_top_features_plot / 
     late_integration_top_features_plot) + 
    plot_layout(heights = c(1, 1, 1))
  ) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

### With hierarchical

```{r}
#| label: fig-combined-top-features-plot-with
#| fig-height: 15
#| fig-width: 10

(
  early_integration_top_features_plot /
  intermediate_integration_top_features_plot /
  late_integration_top_features_plot_faceted /
  wrap_elements(full = hierarchical_integration_top_features_horiz)
) +
  plot_layout(heights = c(1, 1, 2, 1.5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

## Without late

```{r}
#| label: fig-combined-top-features-plot-without-lates
#| fig-height: 10
#| fig-width: 10

(
  early_integration_top_features_plot /
  intermediate_integration_top_features_plot /
  wrap_elements(full = hierarchical_integration_top_features_horiz)
) +
  plot_layout(heights = c(1, 1, 1.5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

# Common features between integration methods

Use only comp1 loadings for intermediate since this component explains all variation in NK %.

```{r}
all_feature_importances_cleaned |> 
  filter(integration != "Hierarchical") |> 
  group_by(feature) |> 
  summarise(
    n_integration = n_distinct(integration),
    integration = paste(unique(integration), collapse = ", "),
    omics = first(omics)
  ) |> 
  filter(n_integration > 1) |> 
  arrange(desc(n_integration)) -> common_features_between_integration_methods

common_features_between_integration_methods |> 
  select(!n_integration) |> 
  arrange(omics) |> 
  gt() |> 
  cols_label(
    feature = md("**Feature**"),
    integration = md("**Integration methods**"),
    omics = md("**Omics type**")
  ) |> 
  cols_width(
    feature ~ px(250),
    integration ~ px(250),
    omics ~ px(150)
  ) |> 
  as_raw_html()
```

```{r}
common_features_between_integration_methods
```

# Comparing model performance

```{r}
model_results <- tribble(
  ~Integration,~Model,~Omics_tested,~R2_percent,~RMSE,
  "Early","Random forest regression","All",98.7,0.128,
  "Intermediate","Multi-block sparse PLS","All",78.4,0.175,
  "Intermediate","Multi-branch neural network","All",97.3,0.038,
  "Hierarchical","Random forest regression","scRNA and Olink",74.0,0.144,
  "Late","LASSO regression","LCMS",95.6,0.083,
  "Late","LASSO regression","miRNA",99.2,0.053,
  "Late","LASSO regression","Olink",99.9,0.044,
  "Late","LASSO regression","scRNA",99.9,0.009
) |> 
  mutate(exact_model = paste0(Integration, ": ", Model))
```

```{r}
r2_plot <- model_results |> 
  ggplot(aes(x = reorder(exact_model, R2_percent), y = R2_percent, fill = Omics_tested)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = paste0(R2_percent)),
    position = position_dodge2(width = 0.9, preserve = "single"),
    hjust = -0.1,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    x = "",
    y = "R² (%)"
  ) +
  scale_fill_brewer(palette = "Paired") +
  # make y axis breaks at 0, 25, 50, 100 but give room at end for text
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100), limits = c(0, 110))
```

```{r}
#| fig-height: 4
#| fig-width: 8

rmse_plot <- model_results |> 
  ggplot(aes(x = reorder(exact_model, -RMSE), y = RMSE, fill = Omics_tested)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = paste0(RMSE)),
    position = position_dodge2(width = 0.9, preserve = "single"),
    hjust = -0.1,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    x = "",
    y = "RMSE"
  ) +
  scale_fill_brewer(palette = "Paired") +
  theme(plot.margin = margin(5, 30, 30, 5)) +
  ylim(0, 0.2)
```

```{r}
#| fig-height: 5.5
#| fig-width: 8

model_results |> 
  pivot_longer(
    cols = c(R2_percent, RMSE),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(metric = recode(metric,
    "R2_percent" = "R² (%)",
    "RMSE" = "RMSE"
  )) |>
  ggplot(aes(x = exact_model, y = value, fill = Omics_tested)) +
  facet_grid(~ metric, scales = "free") +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = paste0(value)),
    position = position_dodge2(width = 0.9, preserve = "single"),
    hjust = -0.1,
    size = 3.5,
    colour = "black"
  ) +
  labs(x = "", y = "") +
  scale_fill_brewer(palette = "Paired", name = "Omics tested") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  theme(
    plot.margin = margin(5, 30, 30, 5),
    legend.position = "bottom",
    legend.justification = "center",
    legend.box.just = "center",
    legend.box = "horizontal",
    legend.title = element_text(hjust = 0.5),
    legend.title.position = "top"
  )

```


```{r}
#| fig-height: 10
#| fig-width: 8

# combine with patchwork
r2_plot / rmse_plot +
  plot_layout(guides = "collect") &
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```






