---
title: "Early integration analysis notebook"
author: "Joseph Oddy"
lightbox: auto
execute:
  echo: false
  warning: false
  dev: "ragg_png"
  cache: false
format:
  html:
    theme: flatly
    mermaid:
      theme: neutral
    page-layout: article
    self-contained: true
    code-line-numbers: true
    code-block-bg: true
    code-block-border-left: "#30BBF0"
    code-copy: true
    code-link: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    fig-align: center
    fig-height: 5
    fig-width: 8
    fig-responsive: true
    fig-dpi: 150
    fig-cap-location: top
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    dev: "ragg_png"
    smooth-scroll: true
    embed-resources: true
---

# Data loading/processing

The first steps in this notebook load the necessary libraries, source the data acquisition script, and process the data. This involves some separate processing of the different data modalities (e.g. scaling) before combining them into a single large matrix with the target variable (% of NK cells in sample) appended to this.

```{r}
#| label: library-loading

library(tidyverse)
library(here)
library(aqua)
library(umap)
library(conflicted)
library(janitor)
library(gt)
library(ggrepel)
library(kernlab)
library(feather)
library(arrow)
library(patchwork)
library(tidymodels)
library(CGTCanalyse)
library(vip)
library(rlang)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::desc)
conflicts_prefer(aqua::rescale)
conflicts_prefer(ComplexUpset::upset)
conflicts_prefer(vip::vip)
conflicted::conflicts_prefer(purrr::map)
conflicts_prefer(base::as.factor)

#load(here("00_data/early_integration_NK_percentage_modelling_environment.RData"))
```

```{r}
#| label: sourcing-data-acquisition-script
#| output: FALSE
#| cache-lazy: FALSE

source(here("01_scripts/data_acquisition.R"))

BPPARAM <- BiocParallel::SnowParam(workers = parallel::detectCores() - 1)
```

```{r}
#| label: scRNA-data-loading

var_features <- setdiff(names(combined_run3_run4_variable_genes_union_df)[-c(1:47)],
                              c("plasticell_sample_id", "label", "omics", "site"))

scRNA_data_long <- combined_pseudobulk_df |>
  select(label, omics, all_of(var_features)) |>
  pivot_longer(
    cols = all_of(var_features),
    names_to = "gene",
    values_to = "normalised_expression"
  ) |>
  group_by(gene) |>
  mutate(scaled_expression = scale(normalised_expression)) |>
  mutate(
    vessel = as.numeric(str_extract(label, "(?<=v)\\d+")),
    day = as.numeric(str_extract(label, "(?<=d)\\d+")),
    run = as.numeric(str_extract(label, "(?<=r)\\d+"))
  ) |>
  relocate(run, vessel, day)

scRNA_data_wide <- scRNA_data_long |>
  select(-normalised_expression) |>
  pivot_wider(
    names_from = gene,
    values_from = scaled_expression
  )
```

```{r}
#| label: lcms-data-loading

lcms_data <- processed_lcms$both_runs_data_wide |>
  mutate(omics = "lcms") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )
```

```{r}
#| label: olink-data-loading

long_olink <- processed_olink$data |>
  filter(!Timepoint == "Fresh Media") |>
  filter(DF == "Neat") |>
  select(label, omics, NPX, Assay, vessel, stream, Timepoint, DF) |>
  filter(Assay %in% processed_olink$m_assays) |>
  group_by(Assay) |>
  mutate(scaled_NPX = scale(NPX)) |>
  ungroup()

wide_olink <- long_olink |>
  select(-NPX) |>
  pivot_wider(
    names_from = Assay,
    values_from = scaled_NPX,
    values_fn = mean
  ) |>
  mutate(
    day = as.numeric(str_remove(Timepoint, "Day ")),
    vessel = as.numeric(str_remove(vessel, "V")),
    run = 4
  ) |>
  relocate(run, day)
```


```{r}
#| label: miRNA-data-loading

miRNA_data <- processed_miRNA$processed_miRNA |>
  mutate(omics = "miRNA") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )

processed_miRNA_long <- miRNA_data |>
  pivot_longer(
    cols = starts_with("ENS"),
    names_to = "miRNA_name",
    values_to = "deseq_normalised_and_vst_expression"
  ) |>
  group_by(miRNA_name) |>
  mutate(scaled_expression = scale(deseq_normalised_and_vst_expression)) |>
  ungroup()

miRNA_data_scaled <- processed_miRNA_long |>
  select(-deseq_normalised_and_vst_expression) |>
  pivot_wider(
    names_from = miRNA_name,
    values_from = scaled_expression
  )
```

```{r}
#| label: process-data-loading

process_data <- processed_process_data$all_process_data_unscaled_added_metrics
```

```{r}
#| label: combining-data

combined_data <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label", suffix = c("_lcms", "_miRNA")
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label", suffix = c("_miRNA", "_olink")
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label", suffix = c("_olink", "_scRNA")
  )

combined_data_names <- names(combined_data) |> 
  as_tibble()

combined_metadata <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-no-missing-data

common_samples <- combined_data |>
  drop_na() |>
  select(label)

lcms_matrix <- lcms_data |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix <- miRNA_data_scaled |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix <- wide_olink |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix <- scRNA_data_wide |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list <- list(
  lcms = lcms_matrix,
  miRNA = miRNA_matrix,
  olink = olink_matrix,
  scRNA = scRNA_matrix
)

stage <- combined_metadata |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  pull(stage)
```

```{r}
#| label: combining-data-and-process-data

combined_data_and_process <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  ) |>
  full_join(
    process_data |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata_and_process <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    process_data |>
      rename(process = "omics") |>
      select(process, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    ifelse(!is.na(process), process, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-with-missing-data
#| cache-lazy: false

summary_metric_samples <- combined_data_and_process |>
  filter(!is.na(cancer_cells_killed_per_PSC)) |>
  select(label)

lcms_matrix_process <- lcms_data |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix_process <- miRNA_data_scaled |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix_process <- wide_olink |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix_process <- scRNA_data_wide |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list_process <- list(
  lcms = lcms_matrix_process,
  miRNA = miRNA_matrix_process,
  olink = olink_matrix_process,
  scRNA = scRNA_matrix_process
)

summary_metric <- process_data |>
  filter(label %in% summary_metric_samples$label) |>
  arrange(label) |>
  select(cancer_cells_killed_per_PSC, label) |>
  column_to_rownames("label") |>
  as.matrix()
```

```{r}
#| label: loading-nk-cell-proportion-data

nk_cells_proportions <- read.csv(here("00_data/nk_cell_proportions_run_4.csv")) |>
  mutate(
    vessel_day = as.character(vessel_day),
    vessel = as.numeric(sub("V(\\d+)_D\\d+", "\\1", vessel_day)),
    day = as.numeric(sub("V\\d+_D(\\d+)", "\\1", vessel_day)),
    label = paste0("r4_v", vessel, "_d", day)
  )

nk_cells_proportions_subset <- nk_cells_proportions |>
  filter(label %in% common_samples$label) |>
  arrange(label)

nk_cells_target <- nk_cells_proportions_subset |>
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()
```

```{r}
#| label: nk-proportions-conversion-to-categorical

nk_cells_proportions_category_data <- nk_cells_proportions |>
  mutate(proportion_NK_category = case_when(
    proportion_NK <= 0.25 ~ "< 25% NKs",
    proportion_NK > 0.25 & proportion_NK <= 0.5 ~ "25 - 50% NKs",
    proportion_NK > 0.5 & proportion_NK <= 0.75 ~ "50 - 75% NKs",
    proportion_NK > 0.75 & proportion_NK <= 1 ~ "> 75% NKs",
  )) |>
  mutate(proportion_NK_category = fct_relevel(
    proportion_NK_category,
    "< 25% NKs",
    "25 - 50% NKs",
    "50 - 75% NKs",
    "> 75% NKs"
  )) |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  select(-X)

nk_cells_proportions_category <- nk_cells_proportions_category_data |>
  column_to_rownames("label") |>
  pull(proportion_NK_category)
```

```{r}
#| label: processing-process-data
#| include: false

list2env(processed_process_data, envir = .GlobalEnv)
```

# Nested cross-validation of random forest model

The model used here to test the early integration method is a random forest regression model. Nested cross-validation is used to assess the performance of the model, with an outer loop for performance assessment and an inner loop for hyperparameter tuning.

```{r}
#| label: rf-data-object

rf_data <- combined_data |> 
  drop_na() |> 
  left_join(nk_cells_proportions_subset |> 
      select(label, proportion_NK),
    by = "label") |> 
  select(-label) |> 
  # add grouping variable for proportion_NK for stratification
  mutate(proportion_NK_quartile = ntile(proportion_NK, 4) |> as.factor()) |> 
  relocate(proportion_NK, proportion_NK_quartile)
```


```{r}
#| label: nested-rf-regression

nested_rf_results <- nested_rf_regression(
  data = rf_data,
  target_variable = "proportion_NK",
  seed = 89,
  trees = 100,
  mtry_range = c(1, 15),
  min_n_range = c(2, 20),
  levels = 4,
  outer_folds = 5,
  outer_repeats = 5,
  inner_folds = 3,
  final_tune_folds = 5,
  metric = "rmse",
  strata = "proportion_NK_quartile"
)
```

## Model assessment

Five-fold cross-validation was specified for the "outer fold" (performance) assessment. The results are shown in @tbl-nested-rf-regression-metrics and @fig-random-forest-regression-truth-vs-predicted. The model has a very high R2, indicating that it explains the variance well. The RMSE is not as good, indicating that the estimates are a little off although they still explain the variance well.

::: panel-tabset
### Model metrics - overall
```{r}
#| label: tbl-nested-rf-regression-metrics-overall

nested_rf_results$outer_performance |>
  unnest(cols = fold) |> 
  group_by(.metric) |> 
  summarise(
    mean_estimate = mean(.estimate, na.rm = TRUE),
    median_estimate = median(.estimate, na.rm = TRUE)
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics - Overall")
  )
```


### Model metrics - each fold
```{r}
#| label: tbl-nested-rf-regression-metrics

nested_rf_results$outer_performance |>
  unnest(cols = fold) |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics")
  )
```

### Plotting true vs. predicted values
```{r}
#| label: fig-random-forest-regression-truth-vs-predicted
#| fig-cap: Random forest regression model truth vs. predictions

ggplot(nested_rf_results$outer_performance |>
  unnest(cols = fold),
  aes(x = .metric, y = .estimate)) +
    geom_boxplot() +
  facet_wrap(~ .metric, scales = "free") +
  theme_catapult()
```
:::

## Final model - parameters

```{r}
#| label: tbl-nested-rf-regression-final-model-parameters

nested_rf_results$final_model$fit$fit$spec$args |>
  map(eval_tidy) |>
  {\(x) tibble(parameter = names(x), value = as.character(x))}() |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Final model - selected parameters")
  )
```


## Feature importance

Having assessed the performance of the model, we can train a new model on all the data. We perform another grid-search and cross-validation to select the best hyperparameters from all the data. Feature importances are then extracted from this model that was trained on all the data. Importantly, this model is not used to assess the performance of the model, only for the feature importances.

@fig-feature-importance-top-10 shows the top 10 features with the greatest importance whereas @fig-feature-importance-number-of-features shows the number of features from each modality that have a positive importance. Notably, the modalities with the greatest number of features represented here are the scRNA and O-link, indicating a potential bias due to their greater dimensionality.

::: panel-tabset
### Ranked features
```{r}
feature_importances <- nested_rf_results$final_model |>
  extract_fit_parsnip() |>
  (\(x) x$fit$variable.importance)() |>
  enframe(name = "feature", value = "importance") |>
  arrange(desc(importance))

lcms_features <- lcms_data |>
  select(!omics:vessel) |> 
  names() |> 
  as_tibble() |> 
  rename(feature = "value") |> 
  mutate(omics = "lcms")

miRNA_features <- miRNA_data_scaled |>
  select(!omics:run) |> 
  names() |> 
  as_tibble() |> 
  rename(feature = "value") |> 
  mutate(omics = "miRNA")

olink_features <- wide_olink |>
  select(!run:DF) |> 
  names() |> 
  as_tibble() |> 
  rename(feature = "value") |> 
  mutate(omics = "olink")

scRNA_features <- scRNA_data_wide |>
  select(!run:omics) |> 
  names() |> 
  as_tibble() |> 
  rename(feature = "value") |> 
  mutate(omics = "scRNA")

all_features <- bind_rows(
  lcms_features,
  miRNA_features,
  olink_features,
  scRNA_features
)

feature_importances_annotated <- feature_importances |> 
  left_join(all_features, by = "feature") |> 
  mutate(omics = case_when(
    str_detect(feature, "lcms") ~ "lcms",
    str_detect(feature, "miRNA") ~ "miRNA",
    str_detect(feature, "olink") ~ "olink",
    str_detect(feature, "scRNA") ~ "scRNA",
    TRUE ~ omics
  )) |> 
  mutate(feature = str_remove(feature, "_lcms|_miRNA|_olink|_scRNA"))
```

```{r}
#| label: fig-feature-importance-top-10
#| fig-height: 4
#| fig-width: 8

ggplot(feature_importances_annotated |> 
         arrange(-importance) |> 
         slice_head(n = 10), aes(x = fct_reorder(feature, importance), y = importance, fill = omics)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Feature importance",
    subtitle = "Top 10 features with greatest importance",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d()
```

### Number of features from each modality
```{r}
#| label: fig-feature-importance-number-of-features
#| fig-width: 8
#| fig-height: 3

if(!here("00_data/early_integration_feature_importances.rds") |> 
   file.exists()) {
  write_rds(feature_importances_annotated, here("00_data/early_integration_feature_importances.rds"))
}

feature_importances_annotated |> 
  filter(importance > 0) |> 
  group_by(omics) |> 
  summarise(n_features = n()) |> 
  ggplot(aes(x = reorder(omics, n_features), y = n_features)) +
  geom_col() +
  theme_catapult() +
  coord_flip() +
  geom_text(
    aes(label = n_features),
    hjust = -0.5,
    size = 3.5,
    colour = "black"
  ) +
  labs(
    title = "Number of features with positive importance",
    subtitle = "Early integration - random forest regression model",
    x = "Omics",
    y = "Number of features"
  )
  
```

### Distribution of feature importances
```{r}
ggplot(feature_importances_annotated |> 
         filter(importance > 0),
       aes(x = importance)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ omics, scales = "free_y") +
  theme_catapult()
```

### Distribution of negative importances
```{r}
ggplot(feature_importances_annotated |> 
         filter(importance < 0),
       aes(x = importance)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ omics, scales = "free_y") +
  theme_catapult()
```
:::

