---
title: "Intermediate integration - regression summary"
author: "Joe Oddy"
lightbox: auto
execute:
  echo: false
  warning: false
  dev: "ragg_png"
  cache: false
format:
  html:
    theme: flatly
    mermaid:
      theme: neutral
    page-layout: article
    self-contained: true
    code-line-numbers: true
    code-block-bg: true
    code-block-border-left: "#30BBF0"
    code-copy: true
    code-link: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    fig-align: center
    fig-height: 5
    fig-width: 8
    fig-responsive: true
    fig-dpi: 150
    fig-cap-location: top
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    dev: "ragg_png"
    smooth-scroll: true
    embed-resources: true
---

```{r}
#| label: library-loading

library(here)
library(aqua)
library(umap)
library(mixOmics)
library(conflicted)
library(janitor)
library(gt)
library(ggrepel)
library(feather)
library(tidymodels)
library(BiocParallel)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::desc)
conflicts_prefer(aqua::rescale)
conflicts_prefer(purrr::map)
```

```{r}
#| label: sourcing-data-acquisition-script
#| output: FALSE
#| cache-lazy: FALSE

source(here("01_scripts/data_acquisition.R"))

# BPPARAM <- BiocParallel::SnowParam(workers = parallel::detectCores() - 1)

# BPPARAM <- BiocParallel::SnowParam(workers = 2, type = "SOCK")
# bpstart(BPPARAM)
```

```{r}
#| label: scRNA-data-loading

var_features <- setdiff(names(combined_run3_run4_variable_genes_union_df)[-c(1:47)],
                              c("plasticell_sample_id", "label", "omics", "site"))

scRNA_data_long <- combined_pseudobulk_df |>
  select(label, omics, all_of(var_features)) |>
  pivot_longer(
    cols = all_of(var_features),
    names_to = "gene",
    values_to = "normalised_expression"
  ) |>
  group_by(gene) |>
  mutate(scaled_expression = scale(normalised_expression)) |>
  mutate(
    vessel = as.numeric(str_extract(label, "(?<=v)\\d+")),
    day = as.numeric(str_extract(label, "(?<=d)\\d+")),
    run = as.numeric(str_extract(label, "(?<=r)\\d+"))
  ) |>
  relocate(run, vessel, day)

scRNA_data_wide <- scRNA_data_long |>
  select(-normalised_expression) |>
  pivot_wider(
    names_from = gene,
    values_from = scaled_expression
  )
```

```{r}
#| label: lcms-data-loading

lcms_data <- processed_lcms$both_runs_data_wide |>
  mutate(omics = "lcms") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )
```

```{r}
#| label: olink-data-loading

long_olink <- processed_olink$data |>
  filter(!Timepoint == "Fresh Media") |>
  filter(DF == "Neat") |>
  select(label, omics, NPX, Assay, vessel, stream, Timepoint, DF) |>
  filter(Assay %in% processed_olink$m_assays) |>
  group_by(Assay) |>
  mutate(scaled_NPX = scale(NPX)) |>
  ungroup()

wide_olink <- long_olink |>
  select(-NPX) |>
  pivot_wider(
    names_from = Assay,
    values_from = scaled_NPX,
    values_fn = mean
  ) |>
  mutate(
    day = as.numeric(str_remove(Timepoint, "Day ")),
    vessel = as.numeric(str_remove(vessel, "V")),
    run = 4
  ) |>
  relocate(run, day)
```


```{r}
#| label: miRNA-data-loading

miRNA_data <- processed_miRNA$processed_miRNA |>
  mutate(omics = "miRNA") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )

processed_miRNA_long <- miRNA_data |>
  pivot_longer(
    cols = starts_with("ENS"),
    names_to = "miRNA_name",
    values_to = "deseq_normalised_and_vst_expression"
  ) |>
  group_by(miRNA_name) |>
  mutate(scaled_expression = scale(deseq_normalised_and_vst_expression)) |>
  ungroup()

miRNA_data_scaled <- processed_miRNA_long |>
  select(-deseq_normalised_and_vst_expression) |>
  pivot_wider(
    names_from = miRNA_name,
    values_from = scaled_expression
  )
```

```{r}
#| label: process-data-loading

process_data <- processed_process_data$all_process_data_unscaled_added_metrics
```

```{r}
#| label: combining-data

combined_data <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-no-missing-data

common_samples <- combined_data |>
  drop_na() |>
  select(label)

lcms_matrix <- lcms_data |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix <- miRNA_data_scaled |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix <- wide_olink |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix <- scRNA_data_wide |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list <- list(
  lcms = lcms_matrix,
  miRNA = miRNA_matrix,
  olink = olink_matrix,
  scRNA = scRNA_matrix
)

stage <- combined_metadata |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  pull(stage)
```

```{r}
#| label: combining-data-and-process-data

combined_data_and_process <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  ) |>
  full_join(
    process_data |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata_and_process <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    process_data |>
      rename(process = "omics") |>
      select(process, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    ifelse(!is.na(process), process, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-with-missing-data
#| cache-lazy: false

summary_metric_samples <- combined_data_and_process |>
  filter(!is.na(cancer_cells_killed_per_PSC)) |>
  select(label)

lcms_matrix_process <- lcms_data |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix_process <- miRNA_data_scaled |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix_process <- wide_olink |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix_process <- scRNA_data_wide |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list_process <- list(
  lcms = lcms_matrix_process,
  miRNA = miRNA_matrix_process,
  olink = olink_matrix_process,
  scRNA = scRNA_matrix_process
)

summary_metric <- process_data |>
  filter(label %in% summary_metric_samples$label) |>
  arrange(label) |>
  select(cancer_cells_killed_per_PSC, label) |>
  column_to_rownames("label") |>
  as.matrix()
```

```{r}
#| label: loading-nk-cell-proportion-data

nk_cells_proportions <- read.csv(here("00_data/nk_cell_proportions_run_4.csv")) |>
  mutate(
    vessel_day = as.character(vessel_day),
    vessel = as.numeric(sub("V(\\d+)_D\\d+", "\\1", vessel_day)),
    day = as.numeric(sub("V\\d+_D(\\d+)", "\\1", vessel_day)),
    label = paste0("r4_v", vessel, "_d", day)
  )

nk_cells_proportions_subset <- nk_cells_proportions |>
  filter(label %in% common_samples$label) |>
  arrange(label)

nk_cells_target <- nk_cells_proportions_subset |>
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()
```

```{r}
#| label: nk-proportions-conversion-to-categorical

nk_cells_proportions_category_data <- nk_cells_proportions |>
  mutate(proportion_NK_category = case_when(
    proportion_NK <= 0.25 ~ "< 25% NKs",
    proportion_NK > 0.25 & proportion_NK <= 0.5 ~ "25 - 50% NKs",
    proportion_NK > 0.5 & proportion_NK <= 0.75 ~ "50 - 75% NKs",
    proportion_NK > 0.75 & proportion_NK <= 1 ~ "> 75% NKs",
  )) |>
  mutate(proportion_NK_category = fct_relevel(
    proportion_NK_category,
    "< 25% NKs",
    "25 - 50% NKs",
    "50 - 75% NKs",
    "> 75% NKs"
  )) |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  select(-X)

nk_cells_proportions_category <- nk_cells_proportions_category_data |>
  column_to_rownames("label") |>
  pull(proportion_NK_category)
```

```{r}
#| label: train-test-split

set.seed(89)

nk_cells_proportions_category_data_split <- nk_cells_proportions_category_data |>
  as.data.frame() |> 
  column_to_rownames("label") |> 
  initial_split(
    prop = 0.66,
    strata = proportion_NK_category
  )

nk_cells_proportions_category_train <- 
  nk_cells_proportions_category_data[nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  pull(proportion_NK_category)

nk_cells_proportions_category_test <- 
  nk_cells_proportions_category_data[-nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  pull(proportion_NK_category)

nk_cells_proportions_train <- 
  nk_cells_proportions_category_data[nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()

nk_cells_proportions_test <- 
  nk_cells_proportions_category_data[-nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()

multiomics_list_train <- multiomics_list |>
  map(~ .x[nk_cells_proportions_category_data_split$in_id, ])

multiomics_list_test <- multiomics_list |>
  map(~ .x[-nk_cells_proportions_category_data_split$in_id, ])
```

```{r}
table(nk_cells_proportions_category_train, useNA = "ifany")
```

```{r}
table(nk_cells_proportions_category_test, useNA = "ifany")
```

# Background

This quarto notebook contains a condensed version of the intermediate integration analysis (specifically the regression analysis) used for the Plasticell project.

## Data processing

Data loading and processing is mostly handled by the data_acquisition.R script. Some further data processing occurs in this report (e.g. scaling, selection of only variable scRNA genes, filtering of Olink data).

# Pairwise sPLS models

In the first step, pairwise sparse PLS models are fit between each modality to explore the relationships between them. In these models, we use one modality to explain another and vice versa. The mode for spls is set to canonical: 

"Canonical (mode = "canonical"): X and Y play symmetric roles. While not mathematically equivalent, the method is quite similar to CCA. It is appropriate to use when there is no a priori relationship between X and Y and as a replacement for CCA in very high dimensional contexts (when variables selection is desired). Y is deflated using information from Y."

Correlations between the latent components are generally high (>0.8). This information can be used to inform the design matrix used for the full DIABLO model.

```{r}
#| label: pairwise-pls-models

# generate pairwise PLS models
pls1 <- spls(multiomics_list[["miRNA"]], multiomics_list[["scRNA"]], scale = FALSE, mode = "canonical")
pls2 <- spls(multiomics_list[["miRNA"]], multiomics_list[["lcms"]], scale = FALSE, mode = "canonical")
pls3 <- spls(multiomics_list[["miRNA"]], multiomics_list[["olink"]], scale = FALSE, mode = "canonical")
pls4 <- spls(multiomics_list[["scRNA"]], multiomics_list[["lcms"]], scale = FALSE, mode = "canonical")
pls5 <- spls(multiomics_list[["scRNA"]], multiomics_list[["olink"]], scale = FALSE, mode = "canonical")
pls6 <- spls(multiomics_list[["lcms"]], multiomics_list[["olink"]], scale = FALSE, mode = "canonical")
```

::: {.panel-tabset}
#### Overall correlations - fig
```{r}
#| label: correlations-between-omics

pls1_correlations <- cor(pls1$variates$X, pls1$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("miRNA_", var1),
    var2 = paste("scRNA_", var2)
  )

pls2_correlations <- cor(pls2$variates$X, pls2$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("miRNA_", var1),
    var2 = paste("lcms_", var2)
  )

pls3_correlations <- cor(pls3$variates$X, pls3$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("miRNA_", var1),
    var2 = paste("olink_", var2)
  )

pls4_correlations <- cor(pls4$variates$X, pls4$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("scRNA_", var1),
    var2 = paste("lcms_", var2)
  )

pls5_correlations <- cor(pls5$variates$X, pls5$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("scRNA_", var1),
    var2 = paste("olink_", var2)
  )

pls6_correlations <- cor(pls6$variates$X, pls6$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("lcms_", var1),
    var2 = paste("olink_", var2)
  )

all_pairwise_correlations <- bind_rows(
  pls1_correlations, pls2_correlations,
  pls3_correlations, pls4_correlations,
  pls5_correlations, pls6_correlations
)
```

```{r}
#| label: fig-pairwise-correlations
#| fig-height: 10
#| fig-width: 8

ggplot(
  all_pairwise_correlations |>
    mutate(comparison = paste(var1, var2, sep = "_")),
  aes(
    x = reorder(comparison, value),
    y = value
  )
) +
  geom_col() +
  theme_catapult() +
  coord_flip()
```

#### Overall correlations - table
```{r}
#| label: tbl-pairwise-spls-correlations

all_pairwise_correlations |>
  arrange(-value) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Pairwise correlations")
  )
```
:::

### Design

A model with an arbitrarily high number of components is initially built. The performance of this model with different number of components can then be evaluated to select the optimal distance metric and number of components.

This initial model includes the following:

* Design matrix specifying correlations of 0.8 between modalities.

* The other hyperparameters (number of components and number of features from each omics) were taken from the tuning of the block.splsda model, since the same tuning cannot be implemented on the regression version of the model.

```{r}
#| label: mixomics-splsda-initial-model
#| output: FALSE

diff_splsda_design <- matrix(0.8,
  ncol = length(multiomics_list_train), nrow = length(multiomics_list_train),
  dimnames = list(names(multiomics_list_train), names(multiomics_list_train))
)

diag(diff_splsda_design) <- 0
```

```{r}
list_keepX_proportion_NK <- list(
  lcms = c(5, 15, 5),
  miRNA = c(5, 5, 5),
  olink = c(15, 15, 5),
  scRNA = c(5, 25, 5)
)
```

# Block sPLS for NK cell proportions

The block sPLS model implemented here does not have built-in capacity for tuning of the hyperparameters or assessment of model accuracy. Consequently, I used a block sPLS-DA model to select appropriate hyperparameters (ncomp and number of features to keep in each block), because these classification models do have tuning and performance assessment functions. For classification purposes, the NK_proportion data was encoded into four classes: < 25% NK, 25 - 50% NK, 50 - 75% NK, and > 75% NK. From this, the optimal number of components was 3 (based on mahalanobis distance) and the optimal number of features from each block to keep in each component are shown below in @tbl-number-of-retained-features-proportion-NK.

```{r}
#| label: tbl-number-of-retained-features-proportion-NK

list_keepX_proportion_NK |>
  as.data.frame() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("modality") |>
  as_tibble() |>
  rename(
    comp1 = V1,
    comp2 = V2
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Features retained from each modality")
  )
```

```{r}
#| label: setting-proportion-model
#| output: false

proportion_spls_model <- block.spls(
  X = multiomics_list_train, Y = nk_cells_proportions_train, ncomp = 3,
  keepX = list_keepX_proportion_NK, design = diff_splsda_design,
  near.zero.var = TRUE, scale = FALSE
)
```

::: {.panel-tabset}
#### Features from each omic
```{r}
#| label: tbl-extracting-features-proportion-model

# the features selected to form the first component
proportion_model_miRNA_features_comp1 <- selectVar(proportion_spls_model, block = "miRNA", comp = 1)$miRNA$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "miRNA"
  )

proportion_model_miRNA_features_comp2 <- selectVar(proportion_spls_model, block = "miRNA", comp = 2)$miRNA$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "miRNA"
  )

proportion_model_scRNA_features_comp1 <- selectVar(proportion_spls_model, block = "scRNA", comp = 1)$scRNA$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "scRNA"
  )
proportion_model_scRNA_features_comp2 <- selectVar(proportion_spls_model, block = "scRNA", comp = 2)$scRNA$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "scRNA"
  )

proportion_model_olink_features_comp1 <- selectVar(proportion_spls_model, block = "olink", comp = 1)$olink$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "olink"
  )

proportion_model_olink_features_comp2 <- selectVar(proportion_spls_model, block = "olink", comp = 2)$olink$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "olink"
  )

proportion_model_lcms_features_comp1 <- selectVar(proportion_spls_model, block = "lcms", comp = 1)$lcms$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "lcms"
  )

proportion_model_lcms_features_comp2 <- selectVar(proportion_spls_model, block = "lcms", comp = 2)$lcms$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "lcms"
  )

proportion_model_all_features <- bind_rows(
  proportion_model_miRNA_features_comp1, proportion_model_miRNA_features_comp2,
  proportion_model_scRNA_features_comp1, proportion_model_scRNA_features_comp2,
  proportion_model_olink_features_comp1, proportion_model_olink_features_comp2,
  proportion_model_lcms_features_comp1, proportion_model_lcms_features_comp2
)

proportion_model_all_features |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Features retained from each modality for comps 1 and 2")
  )
```

#### Variable plot
```{r}
#| label: fig-proportion-variables

plotVar(proportion_spls_model,
  var.names = FALSE,
  style = "graphics", legend = TRUE, overlap = FALSE
)
```

#### Loadings
```{r}
#| label: fig-proportion-loadings

plotLoadings(proportion_spls_model, comp = 1, contrib = "max", method = "median")
```

```{r}
#| label: get-loadings-nk-cell-spls

lcms_proportion_model_loadings <- proportion_spls_model[["loadings"]][["lcms"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "lcms")

miRNA_proportion_model_loadings <- proportion_spls_model[["loadings"]][["miRNA"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "miRNA")

olink_proportion_model_loadings <- proportion_spls_model[["loadings"]][["olink"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "olink")

scRNA_proportion_model_loadings <- proportion_spls_model[["loadings"]][["scRNA"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "scRNA")

proportion_model_loadings <- bind_rows(
  lcms_proportion_model_loadings,
  miRNA_proportion_model_loadings,
  olink_proportion_model_loadings,
  scRNA_proportion_model_loadings
) |>
  mutate(feature_omics = paste(feature, omics, sep = "_"))
```

```{r}
#| fig-height: 8

ggplot(
  proportion_model_loadings |>
    group_by(omics) |>
    slice_max(abs(comp1), n = 5),
  # arrange by comp1 value
  aes(x = reorder(feature, abs(comp1)), y = comp1)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~omics, scales = "free") +
  theme_catapult()
```

```{r}
intermediate_integration_features <- proportion_model_loadings |> 
  filter(comp1 != 0 | comp2 != 0 | comp3 != 0)

if(!file.exists(here("00_data/intermediate_integration_features.rds"))){
  write_rds(intermediate_integration_features, here("00_data/intermediate_integration_features.rds"))
}
```

:::

### Final model - samples

::: {.panel-tabset}
#### Separate blocks PLS
```{r}
#| label: fig-mixomics-samples-proportion-model
#| fig-height: 6

plotIndiv(proportion_spls_model)
```

#### Combined PLS
```{r}
#| label: fig-proportion-combined-pls
#| fig-height: 6

# plot the samples using an arrow plot
plotArrow(proportion_spls_model,
  ind.names = FALSE, # reduce visual clutter
  legend = TRUE,
  title = "sPLS comp 1 - 2"
)
```
:::

### Explained variance

```{r}
proportion_spls_model_explained_variance <- proportion_spls_model$prop_expl_var |>
  as.data.frame() |> 
  rownames_to_column("component")
```

```{r}
pred <- predict(proportion_spls_model, newdata = multiomics_list_test)
```

```{r}
# Predicted NK proportions using 1 component
y_pred_comp1 <- pred$AveragedPredict[, 1, 1]

# Using 2 components
y_pred_comp2 <- pred$AveragedPredict[, 1, 2]

# Using 3 components
y_pred_comp3 <- pred$AveragedPredict[, 1, 3]

Y_test <- nk_cells_proportions_test
```

```{r}
# R-squared for 1 component
cor(y_pred_comp1, Y_test)^2

# RMSE
sqrt(mean((y_pred_comp1 - Y_test)^2))

# MAE (optional)
mean(abs(y_pred_comp1 - Y_test))
```

```{r}
# Extract predictions from the last component (i.e., using all 3 components)
y_pred_all <- pred$AveragedPredict[, 1, 3]

# R-squared
R2_all <- cor(y_pred_all, nk_cells_proportions_test)^2
```

```{r}
df_plot <- data.frame(
  Actual = nk_cells_proportions_test,
  Predicted = y_pred_all
)

ggplot(df_plot, aes(x = proportion_NK, y = Predicted)) +
  geom_point() +
  labs(
    title = "Predicted vs Actual NK Proportions",
    x = "Actual NK Proportions",
    y = "Predicted NK Proportions"
  ) +
  theme_catapult()
```

# Implementing nested cross-validation approach

The issue with not being able to tune hyperparameters (ncomp and keepX) for a block.spls model in mixomics is limiting, so maybe explore implementing our own version here.

I attempted to add the block.spls model as a parsnip model, so that it could be used in a tidymodels framework and allow for tuning. However, it can't be added due to the input data type. The block.spls function takes a list of matrices (from each omics layer) as input, whereas the tidymodels framework requires a single tibble/matrix/dataframe as input. 

Therefore we should just use the block.spls function as used in mixomics and perhaps loop over the cross-validation sets instead.

```{r}
#| label: train-test-split-cv

set.seed(89)

cv_data_split <- nk_cells_proportions_category_data |>
  as.data.frame() |> 
  column_to_rownames("label") |> 
  vfold_cv(
    strata = proportion_NK_category,
    v = 5,
    repeats = 10
  )
```

```{r}
#| label: nested-cross-validation-setup

# loop over my train-test splits to perform the analysis I did above for getting performance metrics

results <- list()

for (i in seq_along(cv_data_split$splits)) {
  
  train_data_ids <- analysis(cv_data_split$splits[[i]]) |> 
    rownames_to_column("label") |> 
    pull(label)
  
  test_data_ids <- assessment(cv_data_split$splits[[i]]) |> 
    rownames_to_column("label") |> 
    pull(label)
  
  target_train <- nk_cells_proportions_category_data |>
    filter(label %in% train_data_ids) |>
    column_to_rownames("label") |>
    select(proportion_NK) |>
    as.matrix()

  target_test <- nk_cells_proportions_category_data |>
    filter(label %in% test_data_ids) |>
    column_to_rownames("label") |>
    select(proportion_NK) |>
    as.matrix()
  
  predictors_train <- multiomics_list |>
    map(~ .x[train_data_ids, , drop = FALSE])

  predictors_test <- multiomics_list |>
    map(~ .x[test_data_ids, , drop = FALSE])
  
  # Fit the model on the training set
  proportion_spls_model <- block.spls(
    X = predictors_train, Y = target_train, ncomp = 3,
    keepX = list_keepX_proportion_NK, design = diff_splsda_design,
    near.zero.var = TRUE, scale = FALSE
  )
  
  # Predict on the test set
  pred <- predict(proportion_spls_model, newdata = predictors_test)
  
  # Calculate performance metrics
  y_pred_all <- pred$AveragedPredict[, 1, 3]
  R2_all <- cor(y_pred_all, target_test)^2
  
  # Store the results (e.g., in a list or data frame)
  results[[i]] <- list(
    R2 = R2_all,
    predictions = y_pred_all,
    actuals = target_test
  )
  # You can also calculate other metrics like RMSE, MAE, etc.
  rmse <- sqrt(mean((y_pred_all - target_test)^2))
  mae <- mean(abs(y_pred_all - target_test))
  # Store these metrics as well
  results[[i]]$rmse <- rmse
  results[[i]]$mae <- mae
}

```

```{r}
#| label: nested-cross-validation-results

results_df <- bind_rows(lapply(results, function(x) {
  data.frame(
    R_squared = x$`R2`,
    RMSE = x$rmse,
    MAE = x$mae
  )
}))

results_df |>
  summarise(
    mean_R2 = mean(proportion_NK),
    sd_R2 = sd(proportion_NK),
    mean_RMSE = mean(RMSE),
    sd_RMSE = sd(RMSE),
    mean_MAE = mean(MAE),
    sd_MAE = sd(MAE)
  )
```

```{r}
#| label: fig-cv-accuracy
#| fig-height: 6
#| fig-width: 8

results_df_long <- results_df |>
  pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "value"
  )

ggplot(results_df_long, aes(x = metric, y = value)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  facet_wrap(~metric, scales = "free") +
  labs(
    title = "Distribution of R-squared values from Nested Cross-Validation",
    x = "R-squared",
    y = "Frequency"
  ) +
  theme_catapult()
```



