---
title: "Intermediate integration analysis (MB-sPLS) notebook"
author: "Joe Oddy"
lightbox: auto
execute:
  echo: false
  warning: false
  dev: "ragg_png"
  cache: false
format:
  html:
    theme: flatly
    mermaid:
      theme: neutral
    page-layout: article
    self-contained: true
    code-line-numbers: true
    code-block-bg: true
    code-block-border-left: "#30BBF0"
    code-copy: true
    code-link: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    fig-align: center
    fig-height: 5
    fig-width: 8
    fig-responsive: true
    fig-dpi: 150
    fig-cap-location: top
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    dev: "ragg_png"
    smooth-scroll: true
    embed-resources: true
---

# Data processing

The first steps in this notebook load the necessary libraries, source the data acquisition script, and process the data. This involves some separate processing of the different data modalities (e.g. scaling) before combining them into a single large matrix with the target variable (% of NK cells in sample) appended to this.

```{r}
#| label: library-loading

library(here)
library(aqua)
library(umap)
library(mixOmics)
library(conflicted)
library(janitor)
library(gt)
library(ggrepel)
library(feather)
library(tidymodels)
library(BiocParallel)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::desc)
conflicts_prefer(aqua::rescale)
conflicts_prefer(purrr::map)
```

```{r}
#| label: sourcing-data-acquisition-script
#| output: FALSE
#| cache-lazy: FALSE

source(here("01_scripts/data_acquisition.R"))

# BPPARAM <- BiocParallel::SnowParam(workers = parallel::detectCores() - 1)

# BPPARAM <- BiocParallel::SnowParam(workers = 2, type = "SOCK")
# bpstart(BPPARAM)
```

```{r}
#| label: scRNA-data-loading

var_features <- setdiff(names(combined_run3_run4_variable_genes_union_df)[-c(1:47)],
                              c("plasticell_sample_id", "label", "omics", "site"))

scRNA_data_long <- combined_pseudobulk_df |>
  select(label, omics, all_of(var_features)) |>
  pivot_longer(
    cols = all_of(var_features),
    names_to = "gene",
    values_to = "normalised_expression"
  ) |>
  group_by(gene) |>
  mutate(scaled_expression = scale(normalised_expression)) |>
  mutate(
    vessel = as.numeric(str_extract(label, "(?<=v)\\d+")),
    day = as.numeric(str_extract(label, "(?<=d)\\d+")),
    run = as.numeric(str_extract(label, "(?<=r)\\d+"))
  ) |>
  relocate(run, vessel, day)

scRNA_data_wide <- scRNA_data_long |>
  select(-normalised_expression) |>
  pivot_wider(
    names_from = gene,
    values_from = scaled_expression
  )
```

```{r}
#| label: lcms-data-loading

lcms_data <- processed_lcms$both_runs_data_wide |>
  mutate(omics = "lcms") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )
```

```{r}
#| label: olink-data-loading

long_olink <- processed_olink$data |>
  filter(!Timepoint == "Fresh Media") |>
  filter(DF == "Neat") |>
  select(label, omics, NPX, Assay, vessel, stream, Timepoint, DF) |>
  filter(Assay %in% processed_olink$m_assays) |>
  group_by(Assay) |>
  mutate(scaled_NPX = scale(NPX)) |>
  ungroup()

wide_olink <- long_olink |>
  select(-NPX) |>
  pivot_wider(
    names_from = Assay,
    values_from = scaled_NPX,
    values_fn = mean
  ) |>
  mutate(
    day = as.numeric(str_remove(Timepoint, "Day ")),
    vessel = as.numeric(str_remove(vessel, "V")),
    run = 4
  ) |>
  relocate(run, day)
```


```{r}
#| label: miRNA-data-loading

miRNA_data <- processed_miRNA$processed_miRNA |>
  mutate(omics = "miRNA") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )

processed_miRNA_long <- miRNA_data |>
  pivot_longer(
    cols = starts_with("ENS"),
    names_to = "miRNA_name",
    values_to = "deseq_normalised_and_vst_expression"
  ) |>
  group_by(miRNA_name) |>
  mutate(scaled_expression = scale(deseq_normalised_and_vst_expression)) |>
  ungroup()

miRNA_data_scaled <- processed_miRNA_long |>
  select(-deseq_normalised_and_vst_expression) |>
  pivot_wider(
    names_from = miRNA_name,
    values_from = scaled_expression
  )
```

```{r}
#| label: process-data-loading

process_data <- processed_process_data$all_process_data_unscaled_added_metrics
```

```{r}
#| label: combining-data

combined_data <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |> 
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-no-missing-data

common_samples <- combined_data |>
  drop_na() |>
  select(label)

lcms_matrix <- lcms_data |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix <- miRNA_data_scaled |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix <- wide_olink |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix <- scRNA_data_wide |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list <- list(
  lcms = lcms_matrix,
  miRNA = miRNA_matrix,
  olink = olink_matrix
)

stage <- combined_metadata |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  pull(stage)
```

```{r}
#| label: combining-data-and-process-data

combined_data_and_process <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  ) |>
  full_join(
    process_data |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata_and_process <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    process_data |>
      rename(process = "omics") |>
      select(process, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    ifelse(!is.na(process), process, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-with-missing-data
#| cache-lazy: false

summary_metric_samples <- combined_data_and_process |>
  filter(!is.na(cancer_cells_killed_per_PSC)) |>
  select(label)

lcms_matrix_process <- lcms_data |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix_process <- miRNA_data_scaled |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix_process <- wide_olink |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix_process <- scRNA_data_wide |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list_process <- list(
  lcms = lcms_matrix_process,
  miRNA = miRNA_matrix_process,
  olink = olink_matrix_process
)

summary_metric <- process_data |>
  filter(label %in% summary_metric_samples$label) |>
  arrange(label) |>
  select(cancer_cells_killed_per_PSC, label) |>
  column_to_rownames("label") |>
  as.matrix()
```

```{r}
#| label: loading-nk-cell-proportion-data-run4

nk_cells_proportions <- read.csv(here("00_data/nk_cell_proportions_run_4.csv")) |>
  mutate(
    vessel_day = as.character(vessel_day),
    vessel = as.numeric(sub("V(\\d+)_D\\d+", "\\1", vessel_day)),
    day = as.numeric(sub("V\\d+_D(\\d+)", "\\1", vessel_day)),
    label = paste0("r4_v", vessel, "_d", day)
  )

nk_cells_proportions_subset <- nk_cells_proportions |>
  filter(label %in% common_samples$label) |>
  arrange(label)

nk_cells_target <- nk_cells_proportions_subset |>
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()
```

```{r}
#| label: nk-proportions-conversion-to-categorical

nk_cells_proportions_category_data <- nk_cells_proportions |>
  mutate(proportion_NK_category = case_when(
    proportion_NK <= 0.25 ~ "< 25% NKs",
    proportion_NK > 0.25 & proportion_NK <= 0.5 ~ "25 - 50% NKs",
    proportion_NK > 0.5 & proportion_NK <= 0.75 ~ "50 - 75% NKs",
    proportion_NK > 0.75 & proportion_NK <= 1 ~ "> 75% NKs",
  )) |>
  mutate(proportion_NK_category = fct_relevel(
    proportion_NK_category,
    "< 25% NKs",
    "25 - 50% NKs",
    "50 - 75% NKs",
    "> 75% NKs"
  )) |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  select(-X)

nk_cells_proportions_category <- nk_cells_proportions_category_data |>
  column_to_rownames("label") |>
  pull(proportion_NK_category)
```

```{r}
#| label: train-test-split

set.seed(89)

nk_cells_proportions_category_data_split <- nk_cells_proportions_category_data |>
  as.data.frame() |> 
  column_to_rownames("label") |> 
  initial_split(
    prop = 0.66,
    strata = proportion_NK_category
  )

nk_cells_proportions_category_train <- 
  nk_cells_proportions_category_data[nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  pull(proportion_NK_category)

nk_cells_proportions_category_test <- 
  nk_cells_proportions_category_data[-nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  pull(proportion_NK_category)

nk_cells_proportions_train <- 
  nk_cells_proportions_category_data[nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()

nk_cells_proportions_test <- 
  nk_cells_proportions_category_data[-nk_cells_proportions_category_data_split$in_id, ] |> 
  remove_rownames() |> 
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()

multiomics_list_train <- multiomics_list |>
  map(~ .x[nk_cells_proportions_category_data_split$in_id, ])

multiomics_list_test <- multiomics_list |>
  map(~ .x[-nk_cells_proportions_category_data_split$in_id, ])
```


# Specifying model design/hyperparameters

## Pairwise sPLS models

In the first step, pairwise sparse PLS models are fit between each modality to explore the relationships between them. In these models, we use one modality to explain another and vice versa. The mode for spls is set to canonical: 

"Canonical (mode = "canonical"): X and Y play symmetric roles. While not mathematically equivalent, the method is quite similar to CCA. It is appropriate to use when there is no a priori relationship between X and Y and as a replacement for CCA in very high dimensional contexts (when variables selection is desired). Y is deflated using information from Y."

Correlations between the latent components are generally high (>0.8) (@fig-pairwise-correlations and @tbl-pairwise-spls-correlations). This information can be used to inform the design matrix used for the full DIABLO model.

```{r}
#| label: pairwise-pls-models

# generate pairwise PLS models
pls1 <- spls(multiomics_list[["miRNA"]], multiomics_list[["lcms"]], scale = FALSE, mode = "canonical")
pls2 <- spls(multiomics_list[["miRNA"]], multiomics_list[["olink"]], scale = FALSE, mode = "canonical")
pls3 <- spls(multiomics_list[["lcms"]], multiomics_list[["olink"]], scale = FALSE, mode = "canonical")
```

::: {.panel-tabset}
### Overall correlations
```{r}
#| label: correlations-between-omics

pls1_correlations <- cor(pls1$variates$X, pls1$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("miRNA_", var1),
    var2 = paste("lcms_", var2)
  )

pls2_correlations <- cor(pls2$variates$X, pls2$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("miRNA_", var1),
    var2 = paste("olink_", var2)
  )

pls3_correlations <- cor(pls3$variates$X, pls3$variates$Y) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    names_to = "var2",
    cols = -var1
  ) |>
  mutate(
    var1 = paste("lcms_", var1),
    var2 = paste("olink_", var2)
  )

all_pairwise_correlations <- bind_rows(
  pls1_correlations, pls2_correlations,
  pls3_correlations
)
```

```{r}
#| label: fig-pairwise-correlations
#| fig-height: 10
#| fig-width: 8

ggplot(
  all_pairwise_correlations |>
    mutate(comparison = paste(var1, var2, sep = "_")),
  aes(
    x = reorder(comparison, value),
    y = value
  )
) +
  geom_col() +
  theme_catapult() +
  coord_flip()
```

### Overall correlations - table
```{r}
#| label: tbl-pairwise-spls-correlations

all_pairwise_correlations |>
  arrange(-value) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Pairwise correlations")
  )
```
:::

## Design

The model design was as follows:

* Design matrix specifying correlations of 0.9 between modalities (based on the pairwise sPLS models).

* The other hyperparameters (number of components and number of features from each omics) were arbitrarily chosen by tuning of a block.splsda (classification) model (described in greater detail below).

The block sPLS model implemented here does not have built-in capacity for tuning of the hyperparameters or assessment of model accuracy. Consequently, I used a block sPLS-DA model to select appropriate hyperparameters (ncomp and number of features to keep in each block), because these classification models do have tuning and performance assessment functions. For classification purposes, the NK_proportion data was encoded into four classes: < 25% NK, 25 - 50% NK, 50 - 75% NK, and > 75% NK. From this, the optimal number of components was 3 (based on mahalanobis distance) and the optimal number of features from each block to keep in each component are shown below in @tbl-number-of-retained-features-proportion-NK.

```{r}
#| label: mixomics-splsda-initial-model
#| output: FALSE

diff_splsda_design <- matrix(0.9,
  ncol = length(multiomics_list_train), nrow = length(multiomics_list_train),
  dimnames = list(names(multiomics_list_train), names(multiomics_list_train))
)

diag(diff_splsda_design) <- 0
```

```{r}
list_keepX_proportion_NK <- list(
  lcms = c(5, 15, 5),
  miRNA = c(5, 5, 5),
  olink = c(15, 15, 5)
)
```

```{r}
#| label: tbl-number-of-retained-features-proportion-NK

list_keepX_proportion_NK |>
  as.data.frame() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("modality") |>
  as_tibble() |>
  rename(
    comp1 = V1,
    comp2 = V2
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Features retained from each modality")
  )
```

# Cross-validated sPLS regression model

For this intermediate integration analysis, a simple five-fold cross-validation method (with 10 repeats) was used to assess performance because hyperparameter tuning is not possible with the mixOmics sPLS regression model.

The metrics are displayed in @fig-cv-accuracy.

```{r}
#| label: train-test-split-cv

set.seed(89)

cv_data_split <- nk_cells_proportions_category_data |>
  as.data.frame() |> 
  column_to_rownames("label") |> 
  vfold_cv(
    strata = proportion_NK_category,
    v = 5,
    repeats = 10
  )
```

```{r}
#| label: nested-cross-validation-setup

# loop over my train-test splits to perform the analysis I did above for getting performance metrics

results <- list()

for (i in seq_along(cv_data_split$splits)) {
  
  train_data_ids <- analysis(cv_data_split$splits[[i]]) |> 
    rownames_to_column("label") |> 
    pull(label)
  
  test_data_ids <- assessment(cv_data_split$splits[[i]]) |> 
    rownames_to_column("label") |> 
    pull(label)
  
  target_train <- nk_cells_proportions_category_data |>
    filter(label %in% train_data_ids) |>
    column_to_rownames("label") |>
    select(proportion_NK) |>
    as.matrix()

  target_test <- nk_cells_proportions_category_data |>
    filter(label %in% test_data_ids) |>
    column_to_rownames("label") |>
    select(proportion_NK) |>
    as.matrix()
  
  predictors_train <- multiomics_list |>
    map(~ .x[train_data_ids, , drop = FALSE])

  predictors_test <- multiomics_list |>
    map(~ .x[test_data_ids, , drop = FALSE])
  
  # Fit the model on the training set
  proportion_spls_model <- block.spls(
    X = predictors_train, Y = target_train, ncomp = 3,
    keepX = list_keepX_proportion_NK, design = diff_splsda_design,
    near.zero.var = TRUE, scale = FALSE
  )
  
  # Predict on the test set
  pred <- predict(proportion_spls_model, newdata = predictors_test)
  
  # Calculate performance metrics
  y_pred_all <- pred$AveragedPredict[, 1, 3]
  R2_all <- cor(y_pred_all, target_test)^2
  
  # Store the results (e.g., in a list or data frame)
  results[[i]] <- list(
    R2 = R2_all,
    predictions = y_pred_all,
    actuals = target_test
  )
  # You can also calculate other metrics like RMSE, MAE, etc.
  rmse <- sqrt(mean((y_pred_all - target_test)^2))
  mae <- mean(abs(y_pred_all - target_test))
  # Store these metrics as well
  results[[i]]$rmse <- rmse
  results[[i]]$mae <- mae
}
```

## Performance metrics - table
```{r}
#| label: nested-cross-validation-results

results_df <- bind_rows(lapply(results, function(x) {
  data.frame(
    R_squared = x$`R2`,
    RMSE = x$rmse,
    MAE = x$mae
  )
}))

results_df_summarised <- results_df |>
  summarise(
    mean_R2 = mean(proportion_NK),
    sd_R2 = sd(proportion_NK),
    mean_RMSE = mean(RMSE),
    sd_RMSE = sd(RMSE),
    mean_MAE = mean(MAE),
    sd_MAE = sd(MAE)
  )

results_df_summarised |>
  gt() |> 
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Performance metrics")
  )

if (!file.exists(here("00_data/diablo_summary_metrics_scRNA_removed.rds"))) {
  write_rds(results_df_summarised, here("00_data/diablo_summary_metrics_scRNA_removed.rds"))
}
```

## Performance metrics - figure
```{r}
#| label: fig-cv-accuracy
#| fig-height: 6
#| fig-width: 8

results_df_long <- results_df |>
  pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "value"
  )

ggplot(results_df_long, aes(x = metric, y = value)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  facet_wrap(~metric, scales = "free") +
  labs(
    title = "Distribution of R-squared values from Cross-Validation",
    x = "Metric",
    y = "Value"
  ) +
  theme_catapult()
```


# Feature importances

Having assessed the model using cross-validation, the final model is fit on the whole data set to find the most informative features. These are listed in @tbl-extracting-features-proportion-model and loadings on component 1 are shown in @fig-loadings-nk-cell-spls.

```{r}
#| label: setting-proportion-model
#| output: false

proportion_spls_model <- block.spls(
  X = multiomics_list, Y = nk_cells_target, ncomp = 3,
  keepX = list_keepX_proportion_NK, design = diff_splsda_design,
  near.zero.var = TRUE, scale = FALSE
)
```

::: {.panel-tabset}
## Features from each omic
```{r}
#| label: tbl-extracting-features-proportion-model

# the features selected to form the first component
proportion_model_miRNA_features_comp1 <- selectVar(proportion_spls_model, block = "miRNA", comp = 1)$miRNA$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "miRNA"
  )

proportion_model_miRNA_features_comp2 <- selectVar(proportion_spls_model, block = "miRNA", comp = 2)$miRNA$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "miRNA"
  )

proportion_model_olink_features_comp1 <- selectVar(proportion_spls_model, block = "olink", comp = 1)$olink$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "olink"
  )

proportion_model_olink_features_comp2 <- selectVar(proportion_spls_model, block = "olink", comp = 2)$olink$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "olink"
  )

proportion_model_lcms_features_comp1 <- selectVar(proportion_spls_model, block = "lcms", comp = 1)$lcms$name |>
  as_tibble() |>
  mutate(
    component = 1,
    omics = "lcms"
  )

proportion_model_lcms_features_comp2 <- selectVar(proportion_spls_model, block = "lcms", comp = 2)$lcms$name |>
  as_tibble() |>
  mutate(
    component = 2,
    omics = "lcms"
  )

proportion_model_all_features <- bind_rows(
  proportion_model_miRNA_features_comp1, proportion_model_miRNA_features_comp2,
  proportion_model_olink_features_comp1, proportion_model_olink_features_comp2,
  proportion_model_lcms_features_comp1, proportion_model_lcms_features_comp2
)

proportion_model_all_features |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Features retained from each modality for comps 1 and 2")
  )
```

## Feature importances
```{r}
#| label: get-loadings-nk-cell-spls

lcms_proportion_model_loadings <- proportion_spls_model[["loadings"]][["lcms"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "lcms")

miRNA_proportion_model_loadings <- proportion_spls_model[["loadings"]][["miRNA"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "miRNA")

olink_proportion_model_loadings <- proportion_spls_model[["loadings"]][["olink"]] |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  mutate(omics = "olink")

proportion_model_loadings <- bind_rows(
  lcms_proportion_model_loadings,
  miRNA_proportion_model_loadings,
  olink_proportion_model_loadings
) |>
  mutate(feature_omics = paste(feature, omics, sep = "_"))
```

```{r}
#| label: fig-loadings-nk-cell-spls
#| fig-height: 8

ggplot(
  proportion_model_loadings |>
    group_by(omics) |>
    slice_max(abs(comp1), n = 5),
  # arrange by comp1 value
  aes(x = reorder(feature, abs(comp1)), y = comp1)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~omics, scales = "free") +
  theme_catapult()
```
:::

```{r}
intermediate_integration_features <- proportion_model_loadings |> 
  filter(comp1 != 0 | comp2 != 0 | comp3 != 0)

if(!file.exists(here("00_data/feature_importances/intermediate_integration_features_scRNA_removed.rds"))){
  write_rds(intermediate_integration_features, here("00_data/feature_importances/intermediate_integration_features_scRNA_removed.rds"))
}
```


