---
title: "Late integration analysis notebook"
author: "Joseph Oddy"
lightbox: auto
execute:
  echo: false
  warning: false
  dev: "ragg_png"
  cache: false
format:
  html:
    theme: flatly
    mermaid:
      theme: neutral
    page-layout: article
    self-contained: true
    code-line-numbers: true
    code-block-bg: true
    code-block-border-left: "#30BBF0"
    code-copy: true
    code-link: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    fig-align: center
    fig-height: 5
    fig-width: 8
    fig-responsive: true
    fig-dpi: 150
    fig-cap-location: top
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    dev: "ragg_png"
    smooth-scroll: true
    embed-resources: true
---

# Data processing and modelling approach

The first steps in this notebook are different to those for the early and intermediate integration analyses, because the different omics are analysed separately so we don't need to keep only those samples which have data for all omics.

The same modelling approach was performed for the four different omics - random forest regression assessed by nested cross-validation. Five-fold cross-validation (with repeats) was specified for the "outer fold" (performance) assessment, and three-fold cross validation for the "inner fold" hyperparameter tuning.

A final model was fit on the full data for each omics modality, which was then used to determine feature importances.

```{r}
#| label: library-loading

library(tidyverse)
library(here)
library(aqua)
library(umap)
library(mixOmics)
# mixKernel has quite a few dependencies, see mixKernel installation guide
library(mixKernel)
library(conflicted)
library(janitor)
library(gt)
library(ggrepel)
library(kernlab)
library(feather)
library(UpSetR)
library(ComplexUpset)
library(arrow)
library(patchwork)
library(tidymodels)
library(CGTCanalyse)
library(vip)
library(rlang)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::desc)
conflicts_prefer(aqua::rescale)
conflicts_prefer(ComplexUpset::upset)
conflicts_prefer(vip::vip)
conflicted::conflicts_prefer(purrr::map)
conflicted::conflicts_prefer(tidyr::expand)
conflicts_prefer(base::as.factor)
conflicts_prefer(arrow::write_feather)

#load(here("00_data/early_integration_NK_percentage_modelling_environment.RData"))
```

```{r}
#| label: sourcing-data-acquisition-script
#| output: FALSE
#| cache-lazy: FALSE

source(here("01_scripts/data_acquisition.R"))

BPPARAM <- BiocParallel::SnowParam(workers = parallel::detectCores() - 1)
```

```{r}
#| label: scRNA-data-loading

var_features <- setdiff(names(combined_run3_run4_variable_genes_union_df)[-c(1:47)],
                              c("plasticell_sample_id", "label", "omics", "site"))

scRNA_data_long <- combined_pseudobulk_df |>
  select(label, omics, all_of(var_features)) |>
  pivot_longer(
    cols = all_of(var_features),
    names_to = "gene",
    values_to = "normalised_expression"
  ) |>
  group_by(gene) |>
  mutate(scaled_expression = scale(normalised_expression)) |>
  mutate(
    vessel = as.numeric(str_extract(label, "(?<=v)\\d+")),
    day = as.numeric(str_extract(label, "(?<=d)\\d+")),
    run = as.numeric(str_extract(label, "(?<=r)\\d+"))
  ) |>
  relocate(run, vessel, day)

scRNA_data_wide <- scRNA_data_long |>
  select(-normalised_expression) |>
  pivot_wider(
    names_from = gene,
    values_from = scaled_expression
  )
```

```{r}
#| label: lcms-data-loading

lcms_data <- processed_lcms$both_runs_data_wide |>
  mutate(omics = "lcms") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )
```

```{r}
#| label: olink-data-loading

long_olink <- processed_olink$data |>
  filter(!Timepoint == "Fresh Media") |>
  filter(DF == "Neat") |>
  select(label, omics, NPX, Assay, vessel, stream, Timepoint, DF) |>
  filter(Assay %in% processed_olink$m_assays) |>
  group_by(Assay) |>
  mutate(scaled_NPX = scale(NPX)) |>
  ungroup()

wide_olink <- long_olink |>
  select(-NPX) |>
  pivot_wider(
    names_from = Assay,
    values_from = scaled_NPX,
    values_fn = mean
  ) |>
  mutate(
    day = as.numeric(str_remove(Timepoint, "Day ")),
    vessel = as.numeric(str_remove(vessel, "V")),
    run = 4
  ) |>
  relocate(run, day)
```

```{r}
#| label: miRNA-data-loading

miRNA_data <- processed_miRNA$processed_miRNA |>
  mutate(omics = "miRNA") |>
  relocate(omics) |>
  mutate(
    run = as.numeric(run),
    vessel = as.numeric(vessel)
  )

processed_miRNA_long <- miRNA_data |>
  pivot_longer(
    cols = starts_with("ENS"),
    names_to = "miRNA_name",
    values_to = "deseq_normalised_and_vst_expression"
  ) |>
  group_by(miRNA_name) |>
  mutate(scaled_expression = scale(deseq_normalised_and_vst_expression)) |>
  ungroup()

miRNA_data_scaled <- processed_miRNA_long |>
  select(-deseq_normalised_and_vst_expression) |>
  pivot_wider(
    names_from = miRNA_name,
    values_from = scaled_expression
  )
```

```{r}
#| label: process-data-loading

process_data <- processed_process_data$all_process_data_unscaled_added_metrics
```

```{r}
#| label: combining-data

combined_data <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label", suffix = c("_lcms", "_miRNA")
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label", suffix = c("_miRNA", "_olink")
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label", suffix = c("_olink", "_scRNA")
  )

combined_data_names <- names(combined_data) |> 
  as_tibble()

combined_metadata <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-no-missing-data

common_samples <- combined_data |>
  drop_na() |>
  select(label)

lcms_matrix <- lcms_data |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix <- miRNA_data_scaled |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix <- wide_olink |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix <- scRNA_data_wide |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list <- list(
  lcms = lcms_matrix,
  miRNA = miRNA_matrix,
  olink = olink_matrix,
  scRNA = scRNA_matrix
)

stage <- combined_metadata |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  pull(stage)
```

```{r}
#| label: combining-data-and-process-data

combined_data_and_process <- lcms_data |>
  select(!c(omics, run:vessel)) |>
  full_join(
    miRNA_data_scaled |>
      select(!c(omics, vessel:run)),
    by = "label"
  ) |>
  full_join(
    wide_olink |>
      select(!c(run:day, omics:DF)),
    by = "label"
  ) |>
  full_join(
    scRNA_data_wide |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  ) |>
  full_join(
    process_data |>
      select(!c(run, vessel, day, omics)),
    by = "label"
  )

combined_metadata_and_process <- lcms_data |>
  rename(lcms = "omics") |>
  select(lcms:vessel) |>
  full_join(
    miRNA_data_scaled |>
      rename(miRNA = "omics") |>
      select(miRNA:run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    wide_olink |>
      rename(olink = "omics") |>
      select(run:DF),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    scRNA_data_wide |>
      rename(scRNA = "omics") |>
      select(scRNA, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  full_join(
    process_data |>
      rename(process = "omics") |>
      select(process, label, vessel, day, run),
    by = c("label", "day", "run", "vessel")
  ) |>
  mutate(stage = case_when(
    day >= 1 & day <= 3 ~ "mesoderm_commitment",
    day >= 4 & day <= 6 ~ "hemogenic_endothelium",
    day >= 8 & day <= 12 ~ "early_haem",
    day >= 13 & day <= 15 ~ "late_haem",
    day >= 19 & day <= 26 ~ "early_NK",
    day >= 29 & day <= 36 ~ "late_NK",
    TRUE ~ "undefined"
  )) |>
  mutate(stage = fct_relevel(
    stage,
    "mesoderm_commitment",
    "hemogenic_endothelium",
    "early_haem",
    "late_haem",
    "early_NK",
    "late_NK"
  )) |>
  mutate(omics = paste(
    ifelse(!is.na(lcms), lcms, ""),
    ifelse(!is.na(miRNA), miRNA, ""),
    ifelse(!is.na(olink), olink, ""),
    ifelse(!is.na(scRNA), scRNA, ""),
    ifelse(!is.na(process), process, ""),
    sep = "+"
  )) |>
  mutate(omics = gsub("\\++", "+", omics)) |>
  mutate(omics = gsub("^\\+|\\+$", "", omics))
```

```{r}
#| label: mixomics-data-preparation-with-missing-data
#| cache-lazy: false

summary_metric_samples <- combined_data_and_process |>
  filter(!is.na(cancer_cells_killed_per_PSC)) |>
  select(label)

lcms_matrix_process <- lcms_data |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:vessel) |>
  as.matrix()

miRNA_matrix_process <- miRNA_data_scaled |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!omics:run) |>
  as.matrix()

olink_matrix_process <- wide_olink |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:DF) |>
  as.matrix()

scRNA_matrix_process <- scRNA_data_wide |>
  right_join(
    summary_metric_samples,
    by = "label"
  ) |>
  arrange(label) |>
  column_to_rownames("label") |>
  select(!run:omics) |>
  as.matrix()

multiomics_list_process <- list(
  lcms = lcms_matrix_process,
  miRNA = miRNA_matrix_process,
  olink = olink_matrix_process,
  scRNA = scRNA_matrix_process
)

summary_metric <- process_data |>
  filter(label %in% summary_metric_samples$label) |>
  arrange(label) |>
  select(cancer_cells_killed_per_PSC, label) |>
  column_to_rownames("label") |>
  as.matrix()
```

```{r}
#| label: loading-nk-cell-proportion-data

nk_cells_proportions <- read.csv(here("00_data/nk_cell_proportions_run_4.csv")) |>
  mutate(
    vessel_day = as.character(vessel_day),
    vessel = as.numeric(sub("V(\\d+)_D\\d+", "\\1", vessel_day)),
    day = as.numeric(sub("V\\d+_D(\\d+)", "\\1", vessel_day)),
    label = paste0("r4_v", vessel, "_d", day)
  )

nk_cells_proportions_subset <- nk_cells_proportions |>
  filter(label %in% common_samples$label) |>
  arrange(label)

nk_cells_target <- nk_cells_proportions_subset |>
  column_to_rownames("label") |>
  select(proportion_NK) |>
  as.matrix()
```

```{r}
#| label: nk-proportions-conversion-to-categorical

nk_cells_proportions_category_data <- nk_cells_proportions |>
  mutate(proportion_NK_category = case_when(
    proportion_NK <= 0.25 ~ "< 25% NKs",
    proportion_NK > 0.25 & proportion_NK <= 0.5 ~ "25 - 50% NKs",
    proportion_NK > 0.5 & proportion_NK <= 0.75 ~ "50 - 75% NKs",
    proportion_NK > 0.75 & proportion_NK <= 1 ~ "> 75% NKs",
  )) |>
  mutate(proportion_NK_category = fct_relevel(
    proportion_NK_category,
    "< 25% NKs",
    "25 - 50% NKs",
    "50 - 75% NKs",
    "> 75% NKs"
  )) |>
  filter(label %in% common_samples$label) |>
  arrange(label) |>
  select(-X)

nk_cells_proportions_category <- nk_cells_proportions_category_data |>
  column_to_rownames("label") |>
  pull(proportion_NK_category)
```

```{r}
#| label: processing-process-data
#| include: false

list2env(processed_process_data, envir = .GlobalEnv)
```

# scRNA modelling

Overall model metrics are shown in @tbl-nested-rf-regression-metrics-overall and predicted vs. true values are shown in @fig-random-forest-regression-truth-vs-predicted. Feature importances are shown in @fig-scRNA-feature-importance.

```{r}
mlp_scRNA_data <- scRNA_data_wide |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(label, proportion_NK) |> 
  drop_na() |> 
  select(!proportion_NK:omics)

if(!file.exists(here("00_data/late_integration_data/scRNA_modelling_data.feather"))){
  write_feather(mlp_scRNA_data, here("00_data/late_integration_data/scRNA_modelling_data.feather"))
}
```

```{r}
#| label: nested-glmnet

scRNA_data_modelling <- scRNA_data_wide |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  drop_na() |> 
  select(!run:omics) |> 
  mutate(proportion_NK_quartile = ntile(proportion_NK, 4) |> as.factor()) |> 
  relocate(proportion_NK, proportion_NK_quartile)

nested_glmnet_results <- nested_lasso_regression(
  data = scRNA_data_modelling,
  target_variable = "proportion_NK",
  seed = 89,
  penalty_range = c(-4, 0),
  levels = 4,
  outer_folds = 5,
  outer_repeats = 5,
  inner_folds = 3,
  final_tune_folds = 5,
  metric = "rmse",
  strata = "proportion_NK_quartile"
)
```

## Model assessment

::: panel-tabset
### Model metrics - overall

```{r}
#| label: tbl-nested-rf-regression-metrics-overall

nested_glmnet_results$outer_performance |>
  unnest(cols = fold) |> 
  group_by(.metric) |> 
  summarise(
    mean_estimate = mean(.estimate, na.rm = TRUE),
    median_estimate = median(.estimate, na.rm = TRUE)
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics - Overall")
  )
```

### Model metrics - each fold

```{r}
#| label: tbl-nested-rf-regression-metrics

nested_glmnet_results$outer_performance |>
  unnest(cols = fold) |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics")
  )
```

### Plotting true vs. predicted values

```{r}
#| label: fig-random-forest-regression-truth-vs-predicted
#| fig-cap: Random forest regression model truth vs. predictions

ggplot(nested_glmnet_results$outer_performance |>
  unnest(cols = fold),
  aes(x = .metric, y = .estimate)) +
    geom_boxplot() +
  facet_wrap(~ .metric, scales = "free") +
  theme_catapult()
```
:::

## Final model - parameters

```{r}
#| label: tbl-nested-rf-regression-final-model-parameters

nested_glmnet_results$final_model$fit$fit$spec$args |>
  map(eval_tidy) |>
  {\(x) tibble(parameter = names(x), value = as.character(x))}() |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Final model - selected parameters")
  )
```

## Feature importance

```{r}
feature_coefficients <- tidy(nested_glmnet_results$final_model) |> 
  filter(estimate != 0) |> 
  arrange(desc(abs(estimate)))
```

```{r}
#| label: fig-scRNA-feature-importance
#| fig-height: 8
#| fig-width: 8

ggplot(feature_coefficients |> 
         filter(!term == "(Intercept)"),
       aes(x = fct_reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Retained features",
    subtitle = "Features with non-zero coefficients",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d()
```

# O-link modelling

Overall model metrics are shown in @tbl-nested-rf-regression-metrics-overall-olink and predicted vs. true values are shown in @fig-random-forest-regression-truth-vs-predicted-olink. Feature importances are shown in @fig-olink-feature-importance.

```{r}
mlp_olink_data <- wide_olink |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(label, proportion_NK) |> 
  drop_na() |> 
  select(!proportion_NK:DF)

if(!file.exists(here("00_data/late_integration_data/olink_data_modelling.feather"))){
  write_feather(mlp_olink_data, here("00_data/late_integration_data/olink_data_modelling.feather"))
}
```

```{r}
#| label: olink-nested-lasso

olink_data_modelling <- wide_olink |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(proportion_NK) |> 
  drop_na() |> 
  select(!run:DF) |> 
  mutate(proportion_NK_quartile = ntile(proportion_NK, 4) |> as.factor()) |> 
  relocate(proportion_NK, proportion_NK_quartile)

olink_lasso_results <- nested_lasso_regression(
  data = olink_data_modelling,
  target_variable = "proportion_NK",
  seed = 89,
  penalty_range = c(-4, 0),
  levels = 4,
  outer_folds = 5,
  outer_repeats = 10,
  inner_folds = 3,
  final_tune_folds = 5,
  metric = "rmse",
  strata = "proportion_NK_quartile"
)
```

## Model assessment

::: panel-tabset
### Model metrics - overall

```{r}
#| label: tbl-nested-rf-regression-metrics-overall-olink

olink_lasso_results$outer_performance |>
  unnest(cols = fold) |> 
  group_by(.metric) |> 
  summarise(
    mean_estimate = mean(.estimate, na.rm = TRUE),
    median_estimate = median(.estimate, na.rm = TRUE)
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics - Overall")
  )
```

### Model metrics - each fold

```{r}
#| label: tbl-nested-rf-regression-metrics-olink

olink_lasso_results$outer_performance |>
  unnest(cols = fold) |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics")
  )
```

### Plotting true vs. predicted values

```{r}
#| label: fig-random-forest-regression-truth-vs-predicted-olink
#| fig-cap: Random forest regression model truth vs. predictions

ggplot(olink_lasso_results$outer_performance |>
  unnest(cols = fold),
  aes(x = .metric, y = .estimate)) +
    geom_boxplot() +
  facet_wrap(~ .metric, scales = "free") +
  theme_catapult()
```
:::

## Final model - parameters

```{r}
#| label: tbl-nested-rf-regression-final-model-parameters-olink

olink_lasso_results$final_model$fit$fit$spec$args |>
  map(eval_tidy) |>
  {\(x) tibble(parameter = names(x), value = as.character(x))}() |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Final model - selected parameters")
  )
```

## Feature importance

```{r}
olink_feature_coefficients <- tidy(olink_lasso_results$final_model) |> 
  filter(estimate != 0) |> 
  arrange(desc(abs(estimate)))
```

```{r}
#| label: fig-olink-feature-importance
#| fig-height: 4
#| fig-width: 8

ggplot(olink_feature_coefficients |> 
         filter(!term == "(Intercept)"),
       aes(x = fct_reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Retained features",
    subtitle = "Features with non-zero coefficients",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d()
```

# miRNA modelling

Overall model metrics are shown in @tbl-nested-rf-regression-metrics-overall-mirna and predicted vs. true values are shown in @fig-random-forest-regression-truth-vs-predicted-mirna. Feature importances are shown in @fig-mirna-feature-importance.

```{r}
mlp_mirna_data <- miRNA_data_scaled |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(label, proportion_NK) |> 
  drop_na() |> 
  select(!proportion_NK:run)

if(!file.exists(here("00_data/late_integration_data/mirna_data_modelling.feather"))){
  write_feather(mlp_mirna_data, here("00_data/late_integration_data/mirna_data_modelling.feather"))
}
```

```{r}
#| label: mirna-nested-lasso

mirna_data_modelling <- miRNA_data_scaled |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(proportion_NK) |> 
  drop_na() |> 
  select(!omics:run) |> 
  mutate(proportion_NK_quartile = ntile(proportion_NK, 4) |> as.factor()) |> 
  relocate(proportion_NK, proportion_NK_quartile)

mirna_lasso_results <- nested_lasso_regression(
  data = mirna_data_modelling,
  target_variable = "proportion_NK",
  seed = 89,
  penalty_range = c(-4, 0),
  levels = 4,
  outer_folds = 5,
  outer_repeats = 10,
  inner_folds = 3,
  final_tune_folds = 5,
  metric = "rmse",
  strata = "proportion_NK_quartile"
)
```

## Model assessment

::: panel-tabset
### Model metrics - overall

```{r}
#| label: tbl-nested-rf-regression-metrics-overall-mirna

mirna_lasso_results$outer_performance |>
  unnest(cols = fold) |> 
  group_by(.metric) |> 
  summarise(
    mean_estimate = mean(.estimate, na.rm = TRUE),
    median_estimate = median(.estimate, na.rm = TRUE)
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics - Overall")
  )
```

### Model metrics - each fold

```{r}
#| label: tbl-nested-rf-regression-metrics-mirna

mirna_lasso_results$outer_performance |>
  unnest(cols = fold) |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics")
  )
```

### Plotting true vs. predicted values

```{r}
#| label: fig-random-forest-regression-truth-vs-predicted-mirna
#| fig-cap: Random forest regression model truth vs. predictions

ggplot(mirna_lasso_results$outer_performance |>
  unnest(cols = fold),
  aes(x = .metric, y = .estimate)) +
    geom_boxplot() +
  facet_wrap(~ .metric, scales = "free") +
  theme_catapult()
```
:::

## Final model - parameters

```{r}
#| label: tbl-nested-rf-regression-final-model-parameters-mirna

mirna_lasso_results$final_model$fit$fit$spec$args |>
  map(eval_tidy) |>
  {\(x) tibble(parameter = names(x), value = as.character(x))}() |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Final model - selected parameters")
  )
```

## Feature importance


```{r}
mirna_feature_coefficients <- tidy(mirna_lasso_results$final_model) |> 
  filter(estimate != 0) |> 
  arrange(desc(abs(estimate)))
```

```{r}
#| label: fig-mirna-feature-importance
#| fig-height: 4
#| fig-width: 8

ggplot(mirna_feature_coefficients |> 
         filter(!term == "(Intercept)"),
       aes(x = fct_reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Retained features",
    subtitle = "Features with non-zero coefficients",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d()
```

# LCMS modelling

Overall model metrics are shown in @tbl-nested-rf-regression-metrics-overall-lcms and predicted vs. true values are shown in @fig-random-forest-regression-truth-vs-predicted-lcms. Feature importances are shown in @fig-lcms-feature-importance.

```{r}
mlp_lcms_data <- lcms_data |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(label, proportion_NK) |> 
  drop_na() |> 
  select(!proportion_NK:vessel)

if(!file.exists(here("00_data/late_integration_data/lcms_data_modelling.feather"))){
  write_feather(mlp_lcms_data, here("00_data/late_integration_data/lcms_data_modelling.feather"))
}
```


```{r}
#| label: lcms-nested-lasso

lcms_data_modelling <- lcms_data |> 
  left_join(nk_cells_proportions |> 
      select(label, proportion_NK),
    by = "label") |> 
  relocate(proportion_NK) |> 
  drop_na() |> 
  select(!omics:vessel) |> 
  mutate(proportion_NK_quartile = ntile(proportion_NK, 4) |> as.factor()) |> 
  relocate(proportion_NK, proportion_NK_quartile)

lcms_lasso_results <- nested_lasso_regression(
  data = lcms_data_modelling,
  target_variable = "proportion_NK",
  seed = 89,
  penalty_range = c(-4, 0),
  levels = 4,
  outer_folds = 5,
  outer_repeats = 10,
  inner_folds = 3,
  final_tune_folds = 5,
  metric = "rmse",
  strata = "proportion_NK_quartile"
)
```

## Model assessment

::: panel-tabset
### Model metrics - overall

```{r}
#| label: tbl-nested-rf-regression-metrics-overall-lcms

lcms_lasso_results$outer_performance |>
  unnest(cols = fold) |> 
  group_by(.metric) |> 
  summarise(
    mean_estimate = mean(.estimate, na.rm = TRUE),
    median_estimate = median(.estimate, na.rm = TRUE)
  ) |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics - Overall")
  )
```

### Model metrics - each fold

```{r}
#| label: tbl-nested-rf-regression-metrics-lcms

lcms_lasso_results$outer_performance |>
  unnest(cols = fold) |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Nested Random Forest Regression Metrics")
  )
```

### Plotting true vs. predicted values

```{r}
#| label: fig-random-forest-regression-truth-vs-predicted-lcms
#| fig-cap: Random forest regression model truth vs. predictions

ggplot(lcms_lasso_results$outer_performance |>
  unnest(cols = fold),
  aes(x = .metric, y = .estimate)) +
    geom_boxplot() +
  facet_wrap(~ .metric, scales = "free") +
  theme_catapult()
```
:::

## Final model - parameters

```{r}
#| label: tbl-nested-rf-regression-final-model-parameters-lcms

lcms_lasso_results$final_model$fit$fit$spec$args |>
  map(eval_tidy) |>
  {\(x) tibble(parameter = names(x), value = as.character(x))}() |> 
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = FALSE,
    use_page_size_select = TRUE
  ) |>
  tab_header(
    title = md("Final model - selected parameters")
  )
```

## Feature importance

```{r}
lcms_feature_coefficients <- tidy(lcms_lasso_results$final_model) |> 
  filter(estimate != 0) |> 
  arrange(desc(abs(estimate))) |> 
  filter(!term == "(Intercept)")
```

```{r}
#| label: fig-lcms-feature-importance
#| fig-height: 4
#| fig-width: 8

ggplot(lcms_feature_coefficients,
       aes(x = fct_reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  theme_catapult() +
  labs(
    title = "Retained features",
    subtitle = "Features with non-zero coefficients",
    x = "Feature",
    y = "Importance"
  ) +
  scale_fill_catapult_d()
```
